---
title: "Hands-on Exercise 3 part I"
author: "Michelle Chen"
date: "29 Apr 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Programming Interactive Data Visualisation with R

### 3.1 Learning Outcome

Learn how to create interactive data visualisations by using **ggiraph** and **plotlyr** packages.

### 3.2 Getting Started

The following R packages are installed:

-   [**ggiraph**](https://davidgohel.github.io/ggiraph/) for making ‘ggplot’ graphics interactive.

-   [**plotly**](https://plotly.com/r/), R library for plotting interactive statistical graphs.

-   [**DT**](https://rstudio.github.io/DT/) provides an R interface to the JavaScript library [DataTables](https://datatables.net/) that create interactive table on html page.

-   [**tidyverse**](https://www.tidyverse.org/), a family of modern R packages specially designed to support data science, analysis and communication task including creating static statistical graphs.

-   [**patchwork**](https://patchwork.data-imaginist.com/) for combining multiple ggplot2 graphs into one figure.

This code is used to accomplish the above task.

```{r}
pacman::p_load(ggiraph, plotly, 
               patchwork, DT, tidyverse) 
```

### 3.3 Import Data

This exercise uses *Exam_data.csv*. This code below is used to import the required data file.

```{r}
exam_data <- read_csv("../data/Exam_data.csv")
```

### 3.4 Interactive Data Visualisation - ggiraph methods

[ggiraph](https://davidgohel.github.io/ggiraph/index.html) is an html widget and a ggplot 2 extension that allows ggplot graphics to be interactive. The interactive data visualisation is made with [**ggplot geometries**](https://davidgohel.github.io/ggiraph/reference/index.html#section-interactive-geometries) that covers:

1.  **Tooltip**: a column of data-sets that contain tooltips to be displayed when the cursor is pointing over elements.
2.  **Onclick**: a column of data-sets that contain a JavaScript function to be executed when elements are clicked.
3.  **Data_id**: a column of data-sets that contain an id to be associated with elements.

### 3.4.1 Tooltip Effect

This code plots an interactive statistical graph for the SCIENCE subject using **ggiraph** package. First, a ggplot object is created. Next, [`girafe()`](https://davidgohel.github.io/ggiraph/reference/girafe.html) of **ggiraph** is used to create an interactive svg object.

```{r}
p <- ggplot(data=exam_data, 
       aes(x = SCIENCE)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE, 
    binwidth = 1, 
    method = "histodot") +
  scale_y_continuous(NULL, 
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618
)
```

### 3.5 Interactivity

The relevant student ID is displayed when the cursor hovers over a data point of interest.

```{r}
#| echo: false
p <- ggplot(data=exam_data, 
       aes(x = SCIENCE)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE, 
    binwidth = 1, 
    method = "histodot") +
  scale_y_continuous(NULL, 
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618
)
```

### 3.5.1 Display multiple information on tooltip

The content of the tooltip can be customised by including a list object.

```{r}
exam_data$tooltip <- c(paste0(     
  "Name = ", exam_data$ID,         
  "\n Class = ", exam_data$CLASS)) 

p <- ggplot(data=exam_data, 
       aes(x = SCIENCE)) +
  geom_dotplot_interactive(
    aes(tooltip = exam_data$tooltip), 
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(
  ggobj = p,
  width_svg = 8,
  height_svg = 8*0.618
)

```

### 3.6 More interactivity

### 3.6.1 Customise tooltip style

We can use [`opts_tooltip()`](https://davidgohel.github.io/ggiraph/reference/opts_tooltip.html) of **ggiraph** to customise tooltip.

```{r}
tooltip_css <- "background-color:yellow; #<<
font-style:bold; color:blue;" #<<

p <- ggplot(data=exam_data, 
       aes(x = SCIENCE)) +
  geom_dotplot_interactive(              
    aes(tooltip = ID),                   
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(    #<<
    opts_tooltip(    #<<
      css = tooltip_css)) #<<
)                                        
```

### 3.6.2 Display statistics on tooltip

There is also an advanced way to customise tooltip. A function is used to compute 90% interval of the mean. The derived statistics are then displayed in the tooltip.

```{r}
tooltip <- function(y, ymax, accuracy = .01) {
  mean <- scales::number(y, accuracy = accuracy)
  sem <- scales::number(ymax - y, accuracy = accuracy)
  paste("Mean science scores:", mean, "+/-", sem)
}

gg_point <- ggplot(data=exam_data, 
                   aes(x = RACE),
) +
  stat_summary(aes(y = SCIENCE, 
                   tooltip = after_stat(  
                     tooltip(y, ymax))),  
    fun.data = "mean_se", 
    geom = GeomInteractiveCol,  
    fill = "light blue"
  ) +
  stat_summary(aes(y = SCIENCE),
    fun.data = mean_se,
    geom = "errorbar", width = 0.2, size = 0.2
  )

girafe(ggobj = gg_point,
       width_svg = 8,
       height_svg = 8*0.618)
```

### **3.6.3 Hover effect with *data_id* aesthetic**

This is another interactive feature of ggiraph - data_id.

```{r}
p <- ggplot(data=exam_data, 
       aes(x = SCIENCE)) +
  geom_dotplot_interactive(           
    aes(data_id = CLASS),             
    stackgroups = TRUE,               
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618                      
)                                        
```

### **3.6.4 Styling hover effect**

The highlight effect can be changed by css codes.

```{r}
p <- ggplot(data=exam_data, 
       aes(x = SCIENCE)) +
  geom_dotplot_interactive(              
    aes(data_id = CLASS),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(                        
    opts_hover(css = "fill: #202020;"),  
    opts_hover_inv(css = "opacity:0.2;") 
  )                                        
)                                        
```

### **3.6.5 Combining tooltip and hover effect**

```{r}
p <- ggplot(data=exam_data, 
       aes(x = SCIENCE)) +
  geom_dotplot_interactive(              
    aes(tooltip = CLASS, 
        data_id = CLASS),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(                        
    opts_hover(css = "fill: #202020;"),  
    opts_hover_inv(css = "opacity:0.2;") 
  )                                        
)                                        
```

Elements associated with the *data_id* will be highlighted from the cursor. The tooltip will also display the CLASS.

### **3.6.6 Click effect with onclick**

`onclick` argument of ggiraph provides hotlink interactivity on the web.

```{r}
exam_data$onclick <- sprintf("window.open(\"%s\")",
"https://www.moe.gov.sg/schoolfinder?journey=Primary%20school",
as.character(exam_data$ID))

p <- ggplot(data=exam_data, 
       aes(x = SCIENCE)) +
  geom_dotplot_interactive(              
    aes(onclick = onclick),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618)                                        
```

### **3.6.7 Coordinated Multiple Views with ggiraph**

When a data point of one of the dotplot is selected, the corresponding data point on the second plot will be highlighted too. This is achieved via:

1.  Appropriate interactive functions of **ggiraph**.

2.  *patchwork* function of [patchwork](https://patchwork.data-imaginist.com/) package to create the interactive coordinated multiple views.

```{r}
p1 <- ggplot(data=exam_data, 
       aes(x = SCIENCE)) +
  geom_dotplot_interactive(              
    aes(data_id = ID),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +  
  coord_cartesian(xlim=c(0,100)) + 
  scale_y_continuous(NULL,               
                     breaks = NULL)

p2 <- ggplot(data=exam_data, 
       aes(x = ENGLISH)) +
  geom_dotplot_interactive(              
    aes(data_id = ID),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") + 
  coord_cartesian(xlim=c(0,100)) + 
  scale_y_continuous(NULL,               
                     breaks = NULL)

girafe(code = print(p1 + p2), 
       width_svg = 6,
       height_svg = 3,
       options = list(
         opts_hover(css = "fill: #202020;"),
         opts_hover_inv(css = "opacity:0.2;")
         )
       ) 
```

## **3.7 Interactive Data Visualisation - plotly methods**

There are two ways to create interactive graph by using plotly, they are:

1.  via *plot_ly()*, and
2.  via *ggplotly()*

### **3.7.1 Create an interactive scatter plot: plot_ly() method**

::: panel-tabset
## The plot

```{r}
#| echo: false 
#| message: false 
#| warning: false

plot_ly(data = exam_data, 
             x = ~SCIENCE, 
             y = ~ENGLISH)
```

## The code

```{r}
plot_ly(data = exam_data, 
             x = ~SCIENCE, 
             y = ~ENGLISH)
```
:::

### **3.7.2 Work with visual variable: plot_ly() method**

::: panel-tabset
## The plot

```{r}
#| echo: false 
#| message: false 
#| warning: false

plot_ly(data = exam_data, 
        x = ~ENGLISH, 
        y = ~MATHS, 
        color = ~RACE)
```

## The code

```{r}
plot_ly(data = exam_data, 
        x = ~ENGLISH, 
        y = ~MATHS, 
        color = ~RACE)
```
:::

### **3.7.3 Create an interactive scatter plot: ggplotly() method**

::: panel-tabset
## The plot

```{r}
#| echo: false 
#| message: false 
#| warning: false

p <- ggplot(data=exam_data, 
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
ggplotly(p)
```

## The code

```{r}
p <- ggplot(data=exam_data, 
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
ggplotly(p)
```
:::

### **3.7.4 Coordinated Multiple Views with plotly()**

The creation of a coordinated linked plot using plotly() involves 3 steps:

1.  [`highlight_key()`](https://www.rdocumentation.org/packages/plotly/versions/4.9.2/topics/highlight_key) of **plotly** package is used as shared data.
2.  scatterplots will be created by using ggplot2 functions.
3.  [*subplot()*](https://plotly.com/r/subplots/) of **plotly** package is used to place them next to each other side-by-side.

::: panel-tabset
## The plot

```{r}
#| echo: false 
#| message: false 
#| warning: false

d <- highlight_key(exam_data)
p1 <- ggplot(data=d, 
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

p2 <- ggplot(data=d, 
            aes(x = MATHS,
                y = SCIENCE)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
subplot(ggplotly(p1),
        ggplotly(p2))
```

## The code

```{r}
d <- highlight_key(exam_data)
p1 <- ggplot(data=d, 
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

p2 <- ggplot(data=d, 
            aes(x = MATHS,
                y = SCIENCE)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
subplot(ggplotly(p1),
        ggplotly(p2))
```
:::

## **3.8 Interactive Data Visualisation - crosstalk methods**

### **3.8.1 Interactive Data Table: DT package**

```{r}
DT::datatable(exam_data, class= "compact")
```

### **3.8.2 Linked brushing: crosstalk method**

::: panel-tabset
## The plot

```{r}
#| echo: false 
#| message: false 
#| warning: false

d <- highlight_key(exam_data) 
p <- ggplot(d, 
            aes(ENGLISH, 
                MATHS)) + 
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

gg <- highlight(ggplotly(p),        
                "plotly_selected")  

crosstalk::bscols(gg,               
                  DT::datatable(d), 
                  widths = 5)        

```

## The code

```{r}
d <- highlight_key(exam_data) 
p <- ggplot(d, 
            aes(ENGLISH, 
                MATHS)) + 
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

gg <- highlight(ggplotly(p),        
                "plotly_selected")  

crosstalk::bscols(gg,               
                  DT::datatable(d), 
                  widths = 5)        
```
:::

## **3.9 Reference**

### **3.9.1 ggiraph**

This [link](https://davidgohel.github.io/ggiraph/index.html) provides online version of the reference guide and several useful articles. Use this [link](https://cran.r-project.org/web/packages/ggiraph/ggiraph.pdf) to download the pdf version of the reference guide.

-   [How to Plot With Ggiraph](https://www.r-bloggers.com/2018/04/how-to-plot-with-ggiraph/)

-   [Interactive map of France with ggiraph](http://rstudio-pubs-static.s3.amazonaws.com/152833_56a4917734204de7b37881d164cf8051.html)

-   [Custom interactive sunbursts with ggplot in R](https://www.pipinghotdata.com/posts/2021-06-01-custom-interactive-sunbursts-with-ggplot-in-r/)

-   This [link](https://github.com/d-qn/2016_08_02_rioOlympicsAthletes) provides code example on how ggiraph is used to interactive graphs for [Swiss Olympians - the solo specialists](https://www.swissinfo.ch/eng/rio-2016-_swiss-olympiansthe-solo-specialists-/42349156?utm_content=bufferd148b&utm_medium=social&utm_source=twitter.com&utm_campaign=buffer).

### **3.9.2 plotly for R**

-   [Getting Started with Plotly in R](https://plotly.com/r/getting-started/)

-   A collection of plotly R graphs are available via this [link](https://plotly.com/r/).

-   Carson Sievert (2020) **Interactive web-based data visualization with R, plotly, and shiny**, Chapman and Hall/CRC is the best resource to learn plotly for R. The online version is available via this [link](https://plotly-r.com/)

-   [Plotly R Figure Reference](https://plotly.com/r/reference/index/) provides a comprehensive discussion of each visual representations.

-   [Plotly R Library Fundamentals](https://plotly.com/r/plotly-fundamentals/) is a good place to learn the fundamental features of Plotly’s R API.

-   [Getting Started](https://gganimate.com/articles/gganimate.html)

-   Visit this [link](https://rpubs.com/raymondteo/dataviz8) for a very interesting implementation of gganimate by your senior.

-   [Building an animation step-by-step with gganimate](https://www.alexcookson.com/post/2020-10-18-building-an-animation-step-by-step-with-gganimate/).

-   [Creating a composite gif with multiple gganimate panels](https://solarchemist.se/2021/08/02/composite-gif-gganimate/)
