---
title: "Hands-on Exercise 2"
author: "Michelle Chen"
date: "23 Apr 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Beyond ggplot2 Fundamentals

### 2.1 Learning Outcome

This exercise introduces ggplot2 extensions to create more elegant and effective statistical graphics. The learning outcomes are:

-   control the placement of annotation on a graph by using functions provided in ggrepel package

-   create professional publication quality figure by using functions provided in ggthemes and hrbrthemes packages

-   plot composite figure by combining ggplot2 graphs by using patchwork package.

### 2.2 Getting started

### 2.2.1 Install and load required libraries

Tidyverse and 4 R packages will be used.

1.  ggrepel: an R package that provides geoms for ggplot2 to repel overlapping text labels.

2.  ggthemes: an R package that provides some extra themes, geoms, and scales for ‘ggplot2’.

3.  hrbrthemes: an R package that provides typography-centric themes and theme components for ggplot2.

4.  patchwork: an R package for preparing composite figure created using ggplot2.

The code chunk below is used to ensure that the necessary packages are loaded into the working R environment.

```{r}
pacman::p_load(ggrepel, patchwork, 
               ggthemes, hrbrthemes,
               tidyverse) 
```

### 2.2.1 Import data

A data file called *Exam_data* will be used. It consists of year end examination grades of a cohort of primary 3 students from a local school. It is in csv file format.

The data is imported using the [*read_csv()*](https://readr.tidyverse.org/reference/read_delim.html) function of [**readr**](https://readr.tidyverse.org/) package.

```{r}
exam_data <- read_csv("../data/Exam_data.csv")
```

### 2.3 Beyond ggplot2 Annotation: ggrepel

As illustrated below, annotation can be a challenge in plotting statistical graphs.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

As a better approach, we use [**ggrepel**](https://ggrepel.slowkow.com/)whichprovides `geoms` for **ggplot2** to repel overlapping text. Hence, replace `geom_text()` by [`geom_text_repel()`](https://ggrepel.slowkow.com/reference/geom_text_repel.html) and `geom_label()` by [`geom_label_repel`](https://ggrepel.slowkow.com/reference/geom_text_repel.html).

### 2.3.1 Working with ggrepel

Using an example of English scores versus Math scores for Primary 3.

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5) +
  geom_label_repel(aes(label = ID), fontface = "bold") +
  coord_cartesian(xlim=c(0,100), ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## The code

```{r}
#| eval: false
#| echo: true
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

Using an example of Science scores versus Math scores for Primary 3.

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(data=exam_data, 
       aes(x= SCIENCE, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5) +
  geom_label_repel(aes(label = ID), fontface = "bold") +
  coord_cartesian(xlim=c(0,100), ylim=c(0,100)) +
  ggtitle("Science scores versus Maths scores for Primary 3")
```

## The code

```{r}
#| eval: false
#| echo: true
ggplot(data=exam_data, 
       aes(x= SCIENCE, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("Science scores versus Maths scores for Primary 3")
```
:::

### 2.4 **Beyond ggplot2 Themes**

ggplot2 comes with eight [built-in themes](https://ggplot2.tidyverse.org/reference/ggtheme.html), they are: `theme_gray()`, `theme_bw()`, `theme_classic()`, `theme_dark()`, `theme_light()`, `theme_linedraw()`, `theme_minimal()`, and `theme_void()`.

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores") 


```

## The code

```{r}
#| eval: false
#| echo: true


ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores") 
```
:::

This [link](https://ggplot2.tidyverse.org/reference/index.html#themes) shares more about ggplot2 `Themes.`

### 2.4.1 **Working with ggtheme package**

[**ggthemes**](https://cran.r-project.org/web/packages/ggthemes/index.html) provides [‘ggplot2’ themes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) that replicate the look of plots by Edward Tufte, Stephen Few, [Fivethirtyeight](https://fivethirtyeight.com/), [The Economist](https://www.economist.com/graphic-detail), ‘Stata’, ‘Excel’, and [The Wall Street Journal](https://www.pinterest.com/wsjgraphics/wsj-graphics/), among others.

In the example below, *The Economist* theme is used.

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()

```

## The code

```{r}
#| eval: false
#| echo: true


ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```
:::

### 2.4.2 **Working with hrbthems package**

[**hrbrthemes**](https://cran.r-project.org/web/packages/hrbrthemes/index.html) package provides a base theme that focuses on typographic elements, including where various labels are placed as well as the fonts that are used.

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()

```

## The code

```{r}
#| eval: false
#| echo: true

ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```
:::

The second goal centers around productivity for a production workflow.

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")

```

## The code

```{r}
#| eval: false
#| echo: true

ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```
:::

::: callout-tip
Learnings

-   `axis_title_size` argument is used to increase the font size of the axis title to 18,

-   `base_size` argument is used to increase the default axis label to 15, and

-   `grid` argument is used to remove the x-axis grid lines.
:::

### 2.5 **Beyond Single Graph**

This section creates composite plot by combining multiple graphs. First, let us create three statistical graphics by using the code chunk below.

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")

```

## The code

```{r}
#| eval: false
#| echo: true

p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")
```
:::

Next:

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")

```

## The code

```{r}
#| eval: false
#| echo: true

p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")
```
:::

Last, we will draw a scatterplot for English score versus Maths score by as shown below.

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## The code

```{r}
#| eval: false
#| echo: true

p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```
:::

### 2.5.1 **Creating Composite Graphics: pathwork methods**

ggplot2 extension - [**patchwork**](https://patchwork.data-imaginist.com/) is specially designed for combining separate ggplot2 graphs into a single figure.

Patchwork package has a very simple syntax where we can create layouts super easily. Here’s the general syntax that combines:

-   Two-Column Layout using the Plus Sign +.

-   Parenthesis () to create a subplot group.

-   Two-Row Layout using the Division Sign `/`

### 2.5.2 **Combining two ggplot2 graphs**

Figure in the tabset below shows a composite of two histograms created using patchwork.

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

p1 + p2
```

## The code

```{r}
#| eval: false
#| echo: true

p1 + p2
```
:::

::: callout-note
The syntax used is very simple.
:::

### 2.5.3 **Combining three ggplot2 graphs**

We can plot more complex composite by using appropriate operators. For example, the composite figure below is plotted by using:

-   “/” operator to stack two ggplot2 graphs,

-   “\|” operator to place the plots beside each other,

-   “()” operator the define the sequence of the plotting.

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

(p1 / p2) | p3
```

## The code

```{r}
#| eval: false
#| echo: true

(p1 / p2) | p3
```
:::

### **2.5.4 Creating a composite figure with tag**

In order to identify subplots in text, **patchwork** also provides auto-tagging capabilities as shown in the figure below.

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')
```

## The code

```{r}
#| eval: false
#| echo: true

((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')
```
:::

### **2.5.5 Creating figure with insert**

Beside providing functions to place plots next to each other based on the provided layout. With [`inset_element()`](https://patchwork.data-imaginist.com/reference/inset_element.html) of **patchwork**, we can place one or several plots or graphic elements freely on top or below another plot.

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```

## The code

```{r}
#| eval: false
#| echo: true

p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```
:::

### **2.5.6 Creating a composite figure by using patchwork and ggtheme**

Figure below is created by combining patchwork and theme_economist() of ggthemes package discussed earlier.

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```

## The code

```{r}
#| eval: false
#| echo: true

patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```
:::

### **2.6 References**

-   Kam, T.S. (2025). [Beyond ggplot2 Fundamentals](https://r4va.netlify.app/chap02).

-   [Patchwork R package goes nerd viral](https://www.littlemissdata.com/blog/patchwork)

-   [ggrepel](https://ggrepel.slowkow.com/)

-   [ggthemes](https://ggplot2.tidyverse.org/reference/ggtheme.html)

-   [hrbrthemes](https://cinc.rud.is/web/packages/hrbrthemes/)

-   [ggplot tips: Arranging plots](https://albert-rapp.de/post/2021-10-28-extend-plot-variety/)

-   [ggplot2 Theme Elements Demonstration](https://henrywang.nl/ggplot2-theme-elements-demonstration/)

-   [ggplot2 Theme Elements Reference Sheet](https://isabella-b.com/blog/ggplot2-theme-elements-reference/)
