---
title: "Hands-on Exercise 5 part V"
author: "Michelle Chen"
date: "14 May 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## **Treemap Visualisation with R**

## **1.1 Overview**

We will be designing treemap using appropriate R packages in the following 3 sections.

1.  Learn how to manipulate transaction data into a treemap strcuture by using selected functions provided in **dplyr** package.
2.  Learn how to plot static treemap by using **treemap** package.
3.  Learn how to design interactive treemap by using **d3treeR** package.

## **1.2 Install and Launch R Packages**

First, **treemap** and **tidyverse** pacakges will need to be installed.

```{r}
pacman::p_load(treemap, treemapify, tidyverse) 
```

## **1.3 Data Wrangling**

*REALIS2018.csv* data, which provides information of private property transaction records in 2018, will be used.

### **1.3.1 Import data set**

```{r}
realis2018 <- read_csv("data/realis2018.csv")
```

The output tibble data.frame is called *realis2018*.

### **1.3.2 Data Wrangling and Manipulation**

The data.frame *realis2018* is in trasaction record form, which is disaggregated and not appropriate to be used to plot a treemap. In this section, we will perform the following steps to manipulate and prepare a data.frame that is appropriate for treemap visualisation:

1.  group transaction records by *Project Name*, *Planning Region*, *Planning Area*, *Property Type* and *Type of Sale*, and
2.  compute *Total Unit Sold*, *Total Area*, *Median Unit Price* and *Median Transacted Price* by applying appropriate summary statistics on *No. of Units*, *Area (sqm)*, *Unit Price (\$ psm)* and *Transacted Price (\$)* respectively.

2 key verbs of **dplyr** package, namely: *group_by()* and *summarize()* will be used to perform these steps.

*group_by()* breaks down a data.frame into specified groups of rows. We then apply the verbs above on the resulting object they’ll be automatically applied “by group”.

Grouping affects the verbs as follows:

- grouped *select()* is the same as ungrouped *select()*, except that grouping variables are always retained.

- grouped *arrange()* is the same as ungrouped; unless you set *.by_group = TRUE*, in which case it orders first by the grouping variables.

- *mutate()* and *filter()* are most useful in conjunction with window functions (like *rank()*, or *min(x) == x*). They are described in detail in vignette(“window-functions”).

- *sample_n()* and *sample_frac()* sample the specified number/fraction of rows in each group.

- *summarise()* computes the summary for each group.

In our case, *group_by()* will be used together with *summarise()* to derive the summarised data.frame.

### **1.3.3 Grouped summaries without the Pipe**

The code below shows a typical two lines code approach to perform the steps.

```{r}
realis2018_grouped <- group_by(realis2018, `Project Name`,
                               `Planning Region`, `Planning Area`, 
                               `Property Type`, `Type of Sale`)
realis2018_summarised <- summarise(realis2018_grouped, 
                          `Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE),
                          `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
                          `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE), 
                          `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))

```

::: callout-note
Aggregation functions such as *sum()* and *median()* obey the usual rule of missing values: if there’s any missing value in the input, the output will be a missing value. The argument *na.rm = TRUE* removes the missing values prior to computation.
:::

The code above is not very efficient because we have to give each intermediate data.frame a name, even though we don’t have to care about it.

### **1.3.4 Grouped summaries with the *pipe***

The code below shows a more efficient way to tackle the same processes by using the *pipe*, %\>%:

```{r}
realis2018_summarised <- realis2018 %>% 
  group_by(`Project Name`,`Planning Region`, 
           `Planning Area`, `Property Type`, 
           `Type of Sale`) %>%
  summarise(`Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE), 
            `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
            `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE),
            `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```

## **1.4 Designing Treemap with treemap Package**

**treemap** package is a R package that offers great flexibility in drawing treemaps. The core function, namely: *treemap()* offers at least 43 arguments. In this section, we will only explore the major arguments for designing treemaps.

### **1.4.1 Designing a static treemap**

We will use *treemap()* to plot a treemap showing the distribution of median unit prices and total unit sold of resale condominium by geographic hierarchy in 2017.

First, we will select records belongs to resale condominium property type from *realis2018_selected* data frame.

```{r}
realis2018_selected <- realis2018_summarised %>%
  filter(`Property Type` == "Condominium", `Type of Sale` == "Resale")
```

### **1.4.2 Using the basic arguments**

The code below designed a treemap by using 3 core arguments of *treemap()*, namely: *index*, *vSize* and *vColor*.

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

::: callout-note
Things to learn from the 3 arguments used:

- index

  - The index vector must consist of at least two column names.

  - If multiple column names are provided, the first name is the highest aggregation level, the second name is the second highest aggregation level, and so on.

- vSize

  - The column must not contain negative values. This is because its vaues will be used to map the sizes of the rectangles of the treemaps.
:::

::: callout-warning
*Warning:*

The treemap above was wrongly coloured. For a correctly designed treemap, the colours of the rectagles should be in different intensity showing, in our case, median unit prices.

For *treemap()*, *vColor* is used in combination with the argument *type* to determine the colours of the rectangles. Without defining *type*, *treemap()* assumes type = index, in our case, the hierarchy of planning areas.
:::

### **1.4.3 Working with *vColor* and *type* arguments**

*type* argument can be defined as follows:

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type = "value",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

::: callout-note
Thinking to learn from the code above.

- The rectangles are coloured with different intensity of green, reflecting their respective median unit prices.

- The legend reveals that the values are binned into ten bins, i.e. 0-5000, 5000-10000, etc. with an equal interval of 5000.
:::

### **1.4.4 Colours in treemap package**

There are 2 arguments that determine the mapping to color palettes: *mapping* and *palette*. The only difference between “value” and “manual” is the default value for mapping. The “value” treemap considers palette to be a diverging color palette, and maps it in such a way that 0 corresponds to the middle color (typically white or yellow), -max(abs(values)) to the left-end color, and max(abs(values)), to the right-end color. The “manual” treemap simply maps min(values) to the left-end color, max(values) to the right-end color, and mean(range(values)) to the middle color.

### **1.4.5 The “value” type treemap**

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="value",
        palette="RdYlBu", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

::: callout-note
Things to learn from the code above:

- Although the colour palette used is RdYlBu, there are no red rectangles in the treemap above. This is because all the median unit prices are positive.

- The reason why we see only 5000 to 45000 in the legend is because the *range* argument is by default (min(values, max(values)) with some pretty rounding.
:::

### **1.4.6 The “manual” type treemap**

The “manual” type does not interpret the values as the “value” type does. Instead, the value range is mapped linearly to the colour palette.

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="RdYlBu", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

::: callout-note
Things to learn from the code above:

- The colour scheme used is very copnfusing. This is because mapping = (min(values), mean(range(values)), max(values)). It is not wise to use diverging colour palette such as *RdYlBu* if the values are all positive or negative
:::

To overcome this problem, a single colour palette such as Blues should be used.

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### **1.4.7 Treemap Layout**

*treemap()* supports two popular treemap layouts:

1.  “squarified” and
2.  “pivotSize” - default.

The squarified treemap algorithm (Bruls et al., 2000) produces good aspect ratios, but ignores the sorting order of the rectangles (sortID). The ordered treemap, pivot-by-size, algorithm (Bederson et al., 2002) takes the sorting order (sortID) into account while aspect ratios are still acceptable.

### **1.4.8 Working with *algorithm* argument**

The code below plots a squarified treemap by changing the *algorithm* argument.

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        algorithm = "squarified",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### **1.4.9 Using *sortID***

When “pivotSize” algorithm is used, *sortID* argument can be used to dertemine the order in which the rectangles are placed from top left to bottom right.

```{r}
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        algorithm = "pivotSize",
        sortID = "Median Transacted Price",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

## **1.5 Designing Treemap using treemapify Package**

**treemapify** is a R package specially developed to draw treemaps in **ggplot2**.

### **1.5.1 Designing a basic treemap**

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`),
       layout = "scol",
       start = "bottomleft") + 
  geom_treemap() +
  scale_fill_gradient(low = "light blue", high = "blue")
```

### **1.5.2 Defining hierarchy**

Group by Planning Region

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`),
       start = "topleft") + 
  geom_treemap()
```

Group by Planning Area

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) + 
  geom_treemap()
```

Adding boundary line

```{r}
ggplot(data=realis2018_selected, 
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) + 
  geom_treemap() +
  geom_treemap_subgroup2_border(colour = "gray40",
                                size = 2) +
  geom_treemap_subgroup_border(colour = "gray20")
```

## 
