---
title: "Hands-on Exercise 8"
author: "Michelle Chen"
date: "28 May 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# **Visualising and Analysing Time-oriented Data**

## 1.1 Learning Outcome

This hands-on exercise focuses on building skillsets for:

1.  plotting a calender heatmap by using ggplot2 functions,
2.  plotting a cycle plot by using ggplot2 functions,
3.  plotting a slopegraph
4.  plotting a horizon chart.

## 1.2 Load Packages

The step below loads the necessarily packages in Rstudio required for this exercise.

```{r}
pacman::p_load(scales, viridis, lubridate, ggthemes,
               gridExtra, readxl, knitr, data.table,
               CGPfunctions, ggHoriPlot, tidyverse)
```

## 1.3 Plot Calendar Heatmap

This section focuses on plotting a calendar heatmap using the ggplot2 package. These are the related learnings:

1.  plot a calender heatmap by using ggplot2 functions and extension,
2.  write function using R programming,
3.  derive specific date and time related field by using base R and lubridate packages
4.  perform data preparation task by using tidyr and dplyr packages.

## 1.3.1 Load the Dataset

The dataset - *eventlog.csv* will be used. It consists of 199,999 rows of time-series cyber attack records by country.

The step below imports the dataset into R and labels the data frame as *attacks*.

```{r}
attacks <- read_csv("data/eventlog.csv")
```

## 1.3.2 Examine the Data Structure

*kable()* is used to review the structure of the imported data frame.

```{r}
kable(head(attacks))
```

There are 3 columns, namely *timestamp*, *source_country* and *tz*.

1.  *timestamp* field stores date-time values in POSIXct format.
2.  *source_country* field stores the source of the attack. It is in *ISO 3166-1 alpha-2* country code.
3.  *tz* field stores time zone of the source IP address.

## 1.3.3 Data Preparation

Step 1: Derive *weekday* and *hour of day* fields.

Two new fields - *wkday* and *hour* need to be derived before plotting the calendar heatmap. The step below is used to perform the task.

```{r}
make_hr_wkday <- function(ts, sc, tz) {
  real_times <- ymd_hms(ts, 
                        tz = tz[1], 
                        quiet = TRUE)
  dt <- data.table(source_country = sc,
                   wkday = weekdays(real_times),
                   hour = hour(real_times))
  return(dt)
  }
```

::: callout-note
- [`ymd_hms()`](https://lubridate.tidyverse.org/reference/ymd_hms.html) and [`hour()`](https://lubridate.tidyverse.org/reference/hour.html) are from [**lubridate**](https://lubridate.tidyverse.org/) package, and

- [`weekdays()`](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/weekdays) is a **base** R function.
:::

Step 2: Derive the attacks tibble data frame.

```{r}
wkday_levels <- c('Saturday', 'Friday', 
                  'Thursday', 'Wednesday', 
                  'Tuesday', 'Monday', 
                  'Sunday')

attacks <- attacks %>%
  group_by(tz) %>%
  do(make_hr_wkday(.$timestamp, 
                   .$source_country, 
                   .$tz)) %>% 
  ungroup() %>% 
  mutate(wkday = factor(
    wkday, levels = wkday_levels),
    hour  = factor(
      hour, levels = 0:23))
```

::: callout-note
`mutate()` of **dplyr** package is used to convert *wkday* and *hour* fields into **factor** so they’ll be ordered when plotting
:::

The table below shows the tidy tibble table post processing.

```{r}
kable(head(attacks))
```

## 1.3.4 Build the Calendar Heatmaps

```{r}
grouped <- attacks %>% 
  count(wkday, hour) %>% 
  ungroup() %>%
  na.omit()

ggplot(grouped, 
       aes(hour, 
           wkday, 
           fill = n)) + 
geom_tile(color = "white", 
          size = 0.1) + 
theme_tufte(base_family = "Helvetica") + 
coord_equal() +
scale_fill_gradient(name = "# of attacks",
                    low = "sky blue", 
                    high = "dark blue") +
labs(x = NULL, 
     y = NULL, 
     title = "Attacks by weekday and time of day") +
theme(axis.ticks = element_blank(),
      plot.title = element_text(hjust = 0.5),
      legend.title = element_text(size = 8),
      legend.text = element_text(size = 6) )
```

::: callout-tip
- a tibble data table called *grouped* is derived by aggregating the attack by *wkday* and *hour* fields.

- a new field called *n* is derived by using `group_by()` and `count()` functions.

- `na.omit()` is used to exclude missing value.

- `geom_tile()` is used to plot tiles (grids) at each x and y position. `color` and `size` arguments are used to specify the border color and line size of the tiles.

- [`theme_tufte()`](https://jrnold.github.io/ggthemes/reference/theme_tufte.html) of [**ggthemes**](https://jrnold.github.io/ggthemes/reference/index.html) package is used to remove unnecessary chart junk. To learn which visual components of default ggplot2 have been excluded, you are encouraged to comment out this line to examine the default plot.

- `coord_equal()` is used to ensure the plot will have an aspect ratio of 1:1.

- `scale_fill_gradient()` function is used to creates a two colour gradient (low-high).
:::

We can group the count by hour and wkday and plot it, there is no need to further preprocess the data since there are values for every combination.

## 1.3.5 Build Multiple Calendar Heatmaps

Next, the focus is to build multiple heatmaps for the top four countries with the highest number of attacks.

Step 1: Derive attack by country object.

In order to identify the required countries, the following needs to be done:

1.  count the number of attacks by country,
2.  calculate the percent of attackes by country, and
3.  save the results in a tibble data frame.

```{r}
attacks_by_country <- count(
  attacks, source_country) %>%
  mutate(percent = percent(n/sum(n))) %>%
  arrange(desc(n))
```

Step 2: Prepare the tidy data frame.

In this step, the focus is on extracting the attack records of the top four countries from *attacks* data frame and save the data in a new tibble data frame (i.e. *top4_attacks*).

```{r}
top4 <- attacks_by_country$source_country[1:4]
top4_attacks <- attacks %>%
  filter(source_country %in% top4) %>%
  count(source_country, wkday, hour) %>%
  ungroup() %>%
  mutate(source_country = factor(
    source_country, levels = top4)) %>%
  na.omit()
```

Step 3: Plot the Multiple Calender Heatmap using ggplot2 package.

```{r}
ggplot(top4_attacks, 
       aes(hour, 
           wkday, 
           fill = n)) + 
  geom_tile(color = "white", 
          size = 0.1) + 
  theme_tufte(base_family = "Helvetica") + 
  coord_equal() +
  scale_fill_gradient(name = "# of attacks",
                    low = "sky blue", 
                    high = "dark blue") +
  facet_wrap(~source_country, ncol = 2) +
  labs(x = NULL, y = NULL, 
     title = "Attacks on top 4 countries by weekday and time of day") +
  theme(axis.ticks = element_blank(),
        axis.text.x = element_text(size = 7),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6) )
```

## 1.3.6 Cycle plot

In this section, the focus is to plot a cycle plot showing the time-series patterns and trend of visitor arrivals from Vietnam programmatically by using ggplot2 functions.

Step 1: Data import.

A different dataset - *arrivals_by_air.xlsx* will be used. The step below imports the dataset using read_excel() of readxl package and save it as a tibble data frame called *air*.

```{r}
air <- read_excel("data/arrivals_by_air.xlsx")
```

Step 2: Derive month and year fields.

Next, two new fields called *month* and *year* are derived from *Month-Year* field.

```{r}
air$month <- factor(month(air$`Month-Year`), 
                    levels=1:12, 
                    labels=month.abb, 
                    ordered=TRUE) 
air$year <- year(ymd(air$`Month-Year`))
```

Step 3: Extract the target country.

Next, the step below is used to extract data for Vietnam.

```{r}
Vietnam <- air %>% 
  select(`Vietnam`, 
         month, 
         year) %>%
  filter(year >= 2010)
```

Step 4: Compute year average arrivals by month.

The step below uses `group_by()` and `summarise()` of **dplyr** to compute year average arrivals by month.

```{r}
hline.data <- Vietnam %>% 
  group_by(month) %>%
  summarise(avgvalue = mean(`Vietnam`))
```

Step 5: Plot the cycle plot.

The step below is used to plot the cycle plot.

```{r}
ggplot() + 
  geom_line(data=Vietnam,
            aes(x=year, 
                y=`Vietnam`, 
                group=month), 
            colour="black") +
  geom_hline(aes(yintercept=avgvalue), 
             data=hline.data, 
             linetype=6, 
             colour="red", 
             size=0.5) + 
  facet_grid(~month) +
  labs(axis.text.x = element_blank(),
       title = "Visitor arrivals from Vietnam by air, Jan 2010-Dec 2019") +
  xlab("") +
  ylab("No. of Visitors") +
  theme_tufte(base_family = "Helvetica")
```

## 1.3.7 Plot Slopegraph

In this section, the focus is to learn how to plot a [slopegraph](https://www.storytellingwithdata.com/blog/2020/7/27/what-is-a-slopegraph) by using R.

Before getting start, **CGPfunctions** needs to be installed and loaded onto R environment.

Step 1: Import data.

The step below imports the required dataset for the slopegraph.

```{r}
rice <- read_csv("data/rice.csv")
```

Step 2: Plot the slopegraph.

The step below is used to plot a basic slopegraph.

```{r}
rice %>% 
  mutate(Year = factor(Year)) %>%
  filter(Year %in% c(1961, 1980)) %>%
  newggslopegraph(Year, Yield, Country,
                Title = "Rice Yield of Top 11 Asian Counties",
                SubTitle = "1961-1980",
                Caption = "Prepared by: Dr. Kam Tin Seong")
```

::: callout-tip
For effective data visualisation design, `factor()` is used convert the value type of *Year* field from numeric to factor.
:::
