---
title: "Hands-on Exercise 5 part III"
author: "Michelle Chen"
date: "13 May 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## **Heatmap for Visualising and Analysing Multivariate Data**

## **1.1 Overview**

Heatmaps visualise data through variations in colouring. When applied to a tabular format, heatmaps are useful for cross-examining multivariate data, through placing variables in the columns and observations in rows and colouring the cells within the table. Heatmaps are good for showing variance across multiple variables, revealing any patterns, displaying whether any variables are similar to each other, and for detecting if any correlations exist in-between them.

We will be focusing on using R to plot static and interactive heatmap for visualising and analysing multivariate data in this exercise.

## **1.2 Install and Launch R Packages**

The code below is used to install and launch **seriation**, **heatmaply**, **dendextend** and **tidyverse** in RStudio.

```{r}
pacman::p_load(seriation, dendextend, heatmaply, tidyverse)
```

## **1.3 Import and Prepare The Data Set**

The data from [World Happines 2018 report](https://worldhappiness.report/ed/2018/) will be used.

### **1.3.1 Import the data set**

```{r}
wh <- read_csv("data/WHData-2018.csv")
```

The output tibbled data frame is called **wh**.

### **1.3.2 Prepare the data**

The rows by number is changed to rows by country name.

```{r}
row.names(wh) <- wh$Country
```

### **1.3.3 Transforming the data frame into a matrix**

The *wh* data frame needs to transform into a data matrix to make a heatmap.

```{r}
wh1 <- dplyr::select(wh, c(3, 7:12))
wh_matrix <- data.matrix(wh)
```

## **1.4 Static Heatmap**

There are many R packages and functions that can be used to draw static heatmaps. They are:

1.  [heatmap()](https://www.rdocumentation.org/packages/stats/versions/3.6.0/topics/heatmap) of R stats package. It draws a simple heatmap.
2.  [heatmap.2()](https://www.rdocumentation.org/packages/gplots/versions/3.0.1.1/topics/heatmap.2) of **gplots** R package. It draws an enhanced heatmap.
3.  [pheatmap()](https://www.rdocumentation.org/packages/pheatmap/versions/1.0.12/topics/pheatmap) of [**pheatmap**](https://www.rdocumentation.org/packages/pheatmap/versions/1.0.12) R package. **pheatmap** package is also known as Pretty Heatmap. The package provides functions to draw pretty heatmaps and provides more control to change the appearance of heatmaps.
4.  [**ComplexHeatmap**](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html) package of R/Bioconductor package. The package draws, annotates and arranges complex heatmaps (very useful for genomic data analysis).
5.  [**superheat**](https://cran.r-project.org/web/packages/superheat/) package. A Graphical Tool for Exploring Complex Datasets Using Heatmaps. A system for generating extendable and customizable heatmaps for exploring complex datasets, including big data and data with multiple data types.

We will be plotting static heatmaps using **heatmap()** of *R Stats* package.

### **1.4.1 heatmap() of R Stats**

```{r}
wh_heatmap <- heatmap(wh_matrix, Rowv=NA, Colv=NA)
```

::: callout-note
By default, **heatmap()** plots a cluster heatmap. The arguments ***Rowv=NA*** and ***Colv=NA*** are used to switch off the option of plotting the row and column dendrograms.
:::

To plot a cluster heatmap, we will use the default as shown.

```{r}
wh_heatmap <- heatmap(wh_matrix)
```

::: callout-note
The order of both rows and columns is different compare to the native wh_matrix. This is because heatmap do a reordering using clusterisation: it calculates the distance between each pair of rows and columns and try to order them by similarity. Moreover, the corresponding dendrogram are provided beside the heatmap.
:::

We need to normalise this matrix as the current heatmap is not informative. The Happiness Score variable has relatively higher values, which makes the other variables with small values all look the same.

```{r}
wh_heatmap <- heatmap(wh_matrix,
                      scale="column",
                      cexRow = 0.6, 
                      cexCol = 0.8,
                      margins = c(10, 4))
```

The values are scaled now. The **margins** argument is used to ensure that the entire x-axis labels are displayed completely and, cexRow and cexCol arguments are used to define the font size used for y-axis and x-axis labels respectively.

## **1.5 Create Interactive Heatmap**

In this section, we will use **heatmaply** to design an interactive cluster heatmap.

### **1.5.1 Working with heatmaply**

```{r}
heatmaply(mtcars)
```

The code below shows the basic syntax needed to create n interactive heatmap by using **heatmaply** package.

```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)])
```

### **1.5.2 Data trasformation**

When analysing multivariate data set, it is very common that the variables in the data sets includes values that reflect different types of measurement. In general, these variables’ values have their own range. In order to ensure that all the variables have comparable values, data transformation are commonly used before clustering.

3 main data transformation methods are supported by *heatmaply()*, namely: scale, normalise and percentise.

#### 1.5.2.1 Scaling method

-   When all variables came from or assumed to come from normal distribution, then scaling (i.e.: subtract the mean and divide by the standard deviation) would bring them all close to the standard normal distribution.

-   Each value would reflect the distance from the mean in units of standard deviation.

-   The *scale* argument in *heatmaply()* supports column and row scaling.

The code below is used to scale variable values columewise.

```{r}
heatmaply(wh_matrix[, -c(1, 2, 4, 5)],
          scale = "column")
```

#### 1.5.2.2 Normalising method

-   When variables in the data comes from different (and non-normal) distributions, the normalize function can be used to bring data to the 0 to 1 scale by subtracting the minimum and dividing by the maximum of all observations.

-   This preserves the shape of each variable’s distribution while making them easily comparable on the same “scale”.

Different from Scaling, the normalise method is performed on the input data set.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]))
```

#### 14.5.2.3 Percentising method

-   This is similar to ranking the variables, but instead of keeping the rank values, divide them by the maximal rank.

-   This is done by using the ecdf of the variables on their own values, bringing each value to its empirical percentile.

-   The benefit of the percentise function is that each value has a relatively clear interpretation, it is the percent of observations that got that value or below it.

Similar to Normalise method, the Percentise method is also performed on the input data set.

```{r}
heatmaply(percentize(wh_matrix[, -c(1, 2, 4, 5)]))
```

### **1.5.3 Clustering algorithm**

**heatmaply** supports a variety of hierarchical clustering algorithm. The main arguments provided are:

-   *distfun*: function used to compute the distance (dissimilarity) between both rows and columns. Defaults to dist. The options “pearson”, “spearman” and “kendall” can be used to use correlation-based clustering, which uses as.dist(1 - cor(t(x))) as the distance metric (using the specified correlation method).

-   *hclustfun*: function used to compute the hierarchical clustering when *Rowv* or *Colv* are not dendrograms. Defaults to *hclust*.

-   *dist_method* default is NULL, which results in “euclidean” to be used. It can accept alternative character strings indicating the method to be passed to distfun. By default *distfun* is “dist”” hence this can be one of “euclidean”, “maximum”, “manhattan”, “canberra”, “binary” or “minkowski”.

-   *hclust_method* default is NULL, which results in “complete” method to be used. It can accept alternative character strings indicating the method to be passed to *hclustfun*. By default hclustfun is hclust hence this can be one of “ward.D”, “ward.D2”, “single”, “complete”, “average” (= UPGMA), “mcquitty” (= WPGMA), “median” (= WPGMC) or “centroid” (= UPGMC).

In general, a clustering model can be calibrated either manually or statistically.

### **1.5.4 Manual approach**

In the code below, the heatmap is plotted by using hierachical clustering algorithm with “Euclidean distance” and “ward.D” method.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "ward.D")
```

### **1.5.5 Statistical approach**

In order to determine the best clustering method and number of cluster the *dend_expend()* and *find_k()* functions of **dendextend** package will be used.

First, the *dend_expend()* will be used to determine the recommended clustering method to be used.

```{r}
wh_d <- dist(normalize(wh_matrix[, -c(1, 2, 4, 5)]), method = "euclidean")
dend_expend(wh_d)[[3]]
```

The output table shows that “average” method should be used because it gave the high optimum value.

Next, *find_k()* is used to determine the optimal number of cluster.

```{r}
wh_clust <- hclust(wh_d, method = "average")
num_k <- find_k(wh_clust)
plot(num_k)
```

Figure above shows that k=3 would be good.

With reference to the statistical analysis results, we can prepare the following.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          dist_method = "euclidean",
          hclust_method = "average",
          k_row = 3)
```

### **1.5.6 Seriation**

One of the problems with hierarchical clustering is that it doesn’t actually place the rows in a definite order, it merely constrains the space of possible orderings.

**heatmaply** uses the seriation package to find an optimal ordering of rows and columns. This means to rotate the branches so that the sum of distances between each adjacent leaf (label) will be minimized.

Here we meet our first seriation algorithm: Optimal Leaf Ordering (OLO). This algorithm starts with the output of an agglomerative clustering algorithm and produces a unique ordering, one that flips the various branches of the dendrogram around so as to minimize the sum of dissimilarities between adjacent leaves. Here is the result of applying Optimal Leaf Ordering to the same clustering result as the heatmap above.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "OLO")
```

Another option is “GW” (Gruvaeus and Wainer) which aims for the same goal but uses a potentially faster heuristic.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "GW")
```

The option “mean” gives the output we would get by default from heatmap functions in other packages such as gplots::heatmap.2.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "mean")
```

The option “none” gives us the dendrograms without any rotation that is based on the data matrix.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "none")
```

### **1.5.7 Working with colour palettes**

The default colour palette uses by **heatmaply** is *viridis*. There are other colour palettes to improve the design aspect of the heatmap.

In the example below, the Blues colour palette of rColorBrewer is used

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          seriate = "none",
          colors = Blues)
```

### **1.5.8 The finishing touch**

*heatmaply* provides many plotting features to ensure cartographic quality heatmap can be produced.

In the code below, the following arguments are used:

-   *k_row* is used to produce 5 groups.

-   *margins* is used to change the top margin to 60 and row margin to 200.

-   *fontsizw_row* and *fontsize_col* are used to change the font size for row and column labels to 4.

-   *main* is used to write the main title of the plot.

-   *xlab* and *ylab* are used to write the x-axis and y-axis labels respectively.

```{r}
heatmaply(normalize(wh_matrix[, -c(1, 2, 4, 5)]),
          Colv=NA,
          seriate = "none",
          colors = Blues,
          k_row = 5,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="World Happiness Score and Variables by Country, 2018 \nDataTransformation using Normalise Method",
          xlab = "World Happiness Indicators",
          ylab = "World Countries"
          )
```
