---
title: "Hands-on Exercise 5 part I"
author: "Michelle Chen"
date: "11 May 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## **Create Ternary Plot with R**

## **1.1 Overview**

Ternary plot is a way to display the distribution and variability of three-part compositional data. Its display is a triangle with sides scaled from 0 to 1. Each side represents one of the three components. A point is plotted so that a line drawn perpendicular from the point to each leg of the triangle intersects at the component values of the point.

We will be building ternary plot using R for visualising and analysing population structure of Singapore.

The hands-on exercise consists of four steps:

1.  Install and launch **tidyverse** and **ggtern** packages.
2.  Derive three new measures using *mutate()* function of **dplyr** package.
3.  Build a static ternary plot using *ggtern()* function of **ggtern** package.
4.  Build an interactive ternary plot using *plot-ly()* function of **Plotly R** package.

## **1.2 Install and launch R packages**

2 main R packages are used in this exercise:

1.  [**ggtern**](http://www.ggtern.com/), a ggplot extension designed to plot ternary diagrams. The package will be used to plot static ternary plots.

- [**Plotly R**](https://plot.ly/r/), an R package for creating interactive web-based graphs via plotly’s JavaScript graphing library, plotly.js.

Selected **tidyverse** family packages - **readr**, **dplyr** and **tidyr** are also required.

The code below loads the required packages.

```{r}
pacman::p_load(plotly, ggtern, tidyverse)
```

## **1.3 Data Preparation**

### **1.3.1 The data**

The [Singapore Residents by Planning AreaSubzone, Age Group, Sex and Type of Dwelling, June 2000-2018](https://www.singstat.gov.sg/find-data/search-by-theme/population/geographic-distribution/latest-data) data will be used.

### **1.3.2 Import Data**

***read_csv()*** function of **readr** package is used to import the required data.

```{r}
pop_data <- read_csv("data/respopagsex2000to2018_tidy.csv")
```

### **1.3 Prepare Data**

***mutate()*** function of **dplyr** package is used to derive 3 new measure - young, active, and old.

```{r}
#Derive the mentioned 3 new measures
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)
```

## **1.4 Plot Ternary Diagram with R**

***ggtern()*** function of **ggtern** package is used to create a simple ternary plot.

```{r}
#Build the static ternary plot
ggtern(data=agpop_mutated,aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point()

```

```{r}
#Build the static ternary plot
ggtern(data=agpop_mutated, aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point() +
  labs(title="Population structure, 2015") +
  theme_rgbw()
```

### **1.4.2 Plot an interative ternary diagram**

The code below creates an interactive ternary plot using ***plot_ly()*** function of **Plotly R**.

```{r}
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# reusable function for axis formatting
axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

# Initiating a plotly visualization 
plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  )
```
