---
title: "Take-home Exercise 1 - Technical Report"
date: "23 May 2026"
date-modified: "last-modified"
format: html
---

# 1 Overview

## 1.1 Background

Motor vehicle insurance consists of heterogeneous customer segments, varying levels of risk exposure, and complex claim behaviours. Insurers rely on data-driven analytical methods to understand the factors influencing portfolio profitability and volatility. Such insights are essential for underwriting, pricing, risk management and strategic business growth decisions.

Traditional reporting approaches for motor vehicle insurance focus on descriptive dashboard summaries. This limits the discovery of hidden relationships and emerging risk patterns within the data. Visual analytics extends beyond static reporting by integrating statistical analysis with exploratory visualisation techniques to uncover meaningful insights from large and complex datasets.

This study applies visual analytics techniques using R, ***tidyverse*** ecosystem, ***ggplot2***, and its extensions to identify key drivers associated with profitability and volatility within a motor vehicle insurance portfolio.

## 1.2 Objectives

The objectives of this study are:

1.  To explore the distributions and characteristics of numerical, categorical, and temporal variables within the insurance portfolio.
2.  To apply visual analytics techniques and investigate factors influencing profitability and volatility.
3.  To statistically validate relationships identified through exploratory visualisations.
4.  To uncover hidden insights that support effective data-driven decision-making within the insurance industry.

## 1.3 Analytical Questions

The study seeks to answer the following key analytical questions:

1.  Which customer and vehicle characteristics are associated with higher insurance profitability?
2.  Which policyholder segments demonstrate higher volatility and risk concentration?
3.  Are there observable temporal trends in premiums, claims, and losses across the portfolio?

## 1.4 The Dataset

The dataset used in this study is a real-world motor vehicle insurance portfolio taken from a Spanish non-life insurance company. It contains **354 140 policy records** across multiple years and includes **47 variables** related to policyholder demographics, vehicle attributes, premium information, claim frequencies, incurred losses, and exposure measures.

The dataset provides a suitable foundation for investigating portfolio profitability and risk volatility. The dataset and its descriptions can be downloaded from [Mendeley Data](https://data.mendeley.com/datasets/sw4jmdb2sm/1).

# 2 Load Packages

## 2.1 Set Analytical Tools

The following packages are required to be loaded in Rstudio before the start of the data analysis.

1.  The ***tidyverse*** family of packages is used extensively for data wrangling and transformation.
2.  ***ggplot2*** and its extensions support advanced visual analytics and exploratory visualisation.
3.  ***scales*** is used for axis formatting and percentage labelling.
4.  ***ggridges*** enables distribution comparison through ridgeline plots.
5.  ***corrplot*** is used to visualise correlation matrix.

The step below is used to load the necessary packages.

```{r}
pacman::p_load(tidyverse, ggplot2, scales, ggridges, corrplot)
```

# 3. Data Preparation and Wrangling

## 3.1 Import Data

The dataset is in csv format and uses semicolons (;) as column delimiters, while using standard periods (.) for decimals. To import the dataset and prevent R from condensing the entire header into a single cell, ***read_delim()*** from the ***readr*** library is used to define the delimiter.

```{r}
#| message: false
#| warning: false

# Use ; to identify separate columns in the csv file
insurance_data <- read_delim(
  "data/Dataset of motor insurance portfolio.csv", 
  delim = ";",
  escape_double = FALSE,
  trim_ws = TRUE
)
```

```{r}
# Verify the number of rows and columns in the csv file
dim(insurance_data)
```

::: callout-note
The original data set has **354140 rows** of records and **47 variables**.
:::

## 3.2 Initial Data Examination

The steps below are used to provide a better understanding of the dataset structure before going into data wrangling decisions.

First, the first 5 records in the dataset is displayed to have an overview of the data types for each variable.

```{r}
head(insurance_data, 5) %>%
  knitr::kable()
```

Next, the general distribution of each variable is generated.

```{r}
summary(insurance_data)
```

## 3.3 Handle Missing Values and Duplicates

The step below identifies the columns with missing values.

```{r}

# Calculate missing values for all columns
missing_counts <- colSums(is.na(insurance_data))
# Display columns that have at least 1 missing value
missing_counts[missing_counts > 0]
```

Out of 47 variables, 4 of them have missing values:

1.  0.0006% of the records for ***vehicle_age*** and ***age_driving_licence*** are missing
2.  0.145% of the records for ***vehicle_value*** are missing
3.  0.363% of the records for ***fuel_type*** are missing

Since the missing data accounts for less than 0.5% of the total observations, list-wise deletion is applied by removing rows containing missing values. Given the extremely small proportion of missing data, the impact on overall analytical interpretation is expected to be minimal.

The step below handles the missing values.

```{r}

# Drop any row that contains an NA in these specific columns
insurance_clean <- insurance_data %>%
  drop_na(vehicle_age, age_driving_licence, fuel_type, vehicle_value)
```

The step below checks for duplicated values.

```{r}

# Count number of rows that are completely duplicated
anyDuplicated(insurance_clean) 
```

There are no duplicated records in the updated ***insurance_clean*** dataset.

## 3.4 Data Type Transformation

There are several variables in the dataset that are recorded in numerical values but should be transformed into categorical data type for more accurate analysis. The step below converts the relevant variables into categorical data types, while ensuring all other columns possess the correct structural attributes for further analysis.

```{r}

insurance_clean <- insurance_clean %>%
  mutate(
# Convert unique ID into text to prevent accidental math operations
    insured_id = as.character(insured_id), 
# Convert the tracking year into categorical factors
    year = as.factor(year),
# Convert all text character columns into categorical factors automatically
    across(
      where(is.character) & -insured_id, as.factor
  )
)
```

::: callout-note
- The unique identifier ***insured_id*** is transformed into a character string to prevent accidental arithmetic calculations.

- The year column is converted into discrete factors to establish independent temporal blocks.

- The ***across()*** function converts the text columns into categorical factors.
:::

## 3.5 Feature Engineering

To deepen the profitability and volatility analysis, several key insurance metrics are derived at the individual policy level. The step below implements the calculations using the ***mutate()*** function.

```{r}

insurance_final <- insurance_clean %>%
  mutate(
    # 1. Net Profit
    net_profit = total_premium - total_incurred,
    
    # 2. Loss Ratio
    loss_ratio = if_else(
      total_premium > 0, total_incurred / total_premium, NA_real_
    ),
    
    # 3. Claim Frequency
    claim_frequency = if_else(
      total_exposure > 0.0, total_claims / total_exposure, NA_real_
    ),
    
    # 4. Claim Severity
    claim_severity = if_else(
      total_claims > 0.0, total_incurred / total_claims, NA_real_
    ),
    
    # 5. Claim indicator 
    has_claim = factor(
      if_else(total_claims > 0.0, "Yes", "No")
    ),
    
    # 6. Driving experience
    driving_experience = driver_age - age_driving_licence,
    experience_group = cut(
      driving_experience,
      breaks = c(0, 5, 10, 20, 30, Inf),
      labels = c("0-5", "6-10", "11-20", "21-30", "30+"),
      include.lowest = TRUE 
    ),
    
    # 7. Vehicle age group 
    vehicle_age_group = cut(
      vehicle_age,
      breaks = c(0, 5, 10, 15, 20, Inf),
      labels = c("0-5", "6-10", "11-15", "16-20", "20+"),
      include.lowest = TRUE  #
    ),
    
    # 8. Driver age group
    age_group = cut(
      driver_age,
      breaks = c(18, 30, 40, 50, 60, Inf),
      labels = c("18-30", "31-40", "41-50", "51-60", "60+"),
      include.lowest = TRUE  
    )
  )

# Preview the clean metrics
#| message: false
#| warning: false

insurance_final %>% 
  select(
    insured_id, net_profit, loss_ratio, claim_frequency, 
    claim_severity, has_claim, experience_group, 
    vehicle_age_group, age_group
  ) %>% 
  head(10) %>%
  knitr::kable()
```

- **Net profit** - calculates whether a policy is profitable or not.
- **Loss ratio** - discovers whether premiums sufficiently cover claims.
- **Claim frequency** - the rate of accidents relative to an active policy.
- **Claim severity** - the average financial cost tied to an individual claim.
- **Claim indicator** - allows claim versus non-claim comparisons.
- **Driving experience in groups**, **Vehicle age in groups**, **Driver age in groups** - are all binned to support easier visual analysis.

```{r}
dim(insurance_final)
```

::: callout-note
The final insurance data set carries **352 338 rows** and **56 columns**.
:::

```{r}
head(insurance_final, 5) %>%
  knitr::kable()
```

# 4. EDA on Portfolio Distribution

This section applies appropriate univariate and bivariate visual analytics techniques to investigate the distributions and characteristics of the motor vehicle insurance portfolio.

## 4.1 Financial Variable Behaviour

Financial variables are explored to understand portfolio profitability, underwriting performance, and the presence of extreme financial observations.

### 4.1.1 Distribution of Total Premium

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(x = total_premium)
) +
  geom_histogram(
    bins = 40,
    fill = "steelblue",
    color = "white"
  ) +
  coord_cartesian(xlim = c(0, 1500)) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Distribution of Total Premium",
    x = "Total Premium",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 14)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(x = total_premium)
) +
  geom_histogram(
    bins = 40,
    fill = "steelblue",
    color = "white"
  ) +
  coord_cartesian(xlim = c(0, 1500)) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Distribution of Total Premium",
    x = "Total Premium",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 14)
```
:::

::: callout-note
The distribution of total premium is right-skewed, indicating that most policyholders contribute relatively low premium amounts, below \$500/year, while a smaller group contributes substantially larger premiums, up to around \$1 250/year.

This suggests that the customer segments have varying levels of risk exposure and coverage requirements.
:::

### 4.1.2 Distribution of Incurred Losses

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(x = log1p(total_incurred))
) +
  geom_histogram(
    bins = 50,
    fill = "steelblue",
    color = "white"
  ) +
  labs(
    title = "Distribution of Total Incurred Losses on Log Scale",
    x = "Log(1 + Total Incurred Losses)",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 14)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(x = log1p(total_incurred))
) +
  geom_histogram(
    bins = 50,
    fill = "steelblue",
    color = "white"
  ) +
  labs(
    title = "Distribution of Total Incurred Losses on Log Scale",
    x = "Log(1 + Total Incurred Losses)",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 14)
```
:::

::: callout-note
Log transformation is applied to improve visual understanding by compressing extreme outliers while preserving the underlying distributional behaviour of insurance claims.

The log-transformed distribution of total incurred losses demonstrates that the portfolio is dominated by policies with little or no financial losses. A relatively small proportion generates more claim events. The heavy concentration near zero indicates that claim occurrence is relatively infrequent, whereas the extended right tail reflects the presence of rare but financially significant losses.
:::

### 4.1.3 Distribution of Claim Severity

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(x = claim_severity)
) +
  geom_histogram(
    bins = 50,
    fill = "steelblue",
    color = "white"
  ) +
  coord_cartesian(
    xlim = c(0, 20000)
  ) +
  scale_x_continuous(
    labels = scales::comma
  ) +
  labs(
    title = "Distribution of Claim Severity",
    x = "Claim Severity",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 14)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(x = claim_severity)
) +
  geom_histogram(
    bins = 50,
    fill = "steelblue",
    color = "white"
  ) +
  coord_cartesian(
    xlim = c(0, 20000)
  ) +
  scale_x_continuous(
    labels = scales::comma
  ) +
  labs(
    title = "Distribution of Claim Severity",
    x = "Claim Severity",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 14)
```
:::

::: callout-note
The claim severity distribution contains extreme outliers, indicating that a small number of claims generate disproportionately high financial losses.
:::

## 4.2 Customer Risk Structure

Customer-related variables are explored to understand the demographic and behavioural characteristics of the insurance portfolio.

### 4.2.1 Distribution of Driver Age Groups

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(x = age_group)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Driver Age Groups",
    x = "Driver Age Group",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(x = age_group)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Driver Age Groups",
    x = "Driver Age Group",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)

```
:::

::: callout-note
Most of the policyholders are in the 41 to 50 years old age group, followed by 31 to 40 years old and 51 to 60 years old. A very small portion of them are 30 years old and below.
:::

### 4.2.2 Distribution of Driving Experience Groups

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(x = experience_group)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Driving Experience Groups",
    x = "Driving Experience Group",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(x = experience_group)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Driving Experience Groups",
    x = "Driving Experience Group",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```
:::

::: callout-note
Most of the policyholders have more than 10 years of driving experience. This aligns with distribution of driver age groups where most of them are above 30 years old.
:::

### 4.2.3 Distribution of Bonus Score

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(x = bonus_score)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Bonus Score",
    x = "Bonus Score",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(x = bonus_score)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Bonus Score",
    x = "Bonus Score",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```
:::

::: callout-note
Most of the policyholders have a positive bonus score which reflects that they have favourable claims history. This suggests that they are generally less risky policyholders.
:::

## 4.3 Vehicle Risk Structure

### 4.3.1 Distribution of Vehicle Value

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(x = vehicle_value)
) +
  geom_histogram(
    bins = 40,
    fill = "steelblue",
    color = "white"
  ) +
  coord_cartesian(xlim = c(0, 50000)) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Distribution of Vehicle Value",
    x = "Vehicle Value",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 14)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(x = vehicle_value)
) +
  geom_histogram(
    bins = 40,
    fill = "steelblue",
    color = "white"
  ) +
  coord_cartesian(xlim = c(0, 50000)) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Distribution of Vehicle Value",
    x = "Vehicle Value",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 14)
```
:::

::: callout-note
The vehicle values are right skewed, with most of them ranging between \$15 000 to \$35 000. This suggests that most insured vehicles are of moderate market values while a smaller segment containing substantially higher-valued vehicles.
:::

### 4.3.2 Distribution of Vehicle Age Groups

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(x = vehicle_age_group)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Vehicle Age Groups",
    x = "Vehicle Age Group",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(x = vehicle_age_group)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Vehicle Age Groups",
    x = "Vehicle Age Group",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```
:::

::: callout-note
Most insured vehicles are 20 years old and above, which may suggests that they have higher maintenance and operational risks.
:::

### 4.3.3 Distribution of Fuel Types

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(x = fuel_type)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Fuel Types",
    x = "Fuel Type",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(x = fuel_type)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Fuel Types",
    x = "Fuel Type",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```
:::

::: callout-note
Most of the vehicles are powered by diesel instead of gasoline.
:::

## 4.4 Portfolio Composition

### 4.4.1 Distribution of Policy Types

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(x = policy_type)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Policy Types",
    x = "Policy Type",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(x = policy_type)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Policy Types",
    x = "Policy Type",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```
:::

::: callout-note
The portfolio composition reveals that third-party liability combined with two or more additional coverage is the most common insurance purchased.
:::

### 4.4.2 Distribution of Claim Occurrence

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(x = has_claim)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Claim Occurrence",
    x = "Claim Indicator",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(x = has_claim)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Claim Occurrence",
    x = "Claim Indicator",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```
:::

::: callout-note
The portfolio is dominated by non-claim policies, which is not surprising in insurance portfolios where claims occur relatively infrequently.
:::

## 4.5 Correlation Matrix

::: panel-tabset
## The matrix

```{r}
#| echo: false
#| message: false
#| warning: false

numeric_vars <- insurance_final %>%
  select(
    total_premium,
    total_incurred,
    net_profit,
    loss_ratio,
    claim_frequency,
    claim_severity,
    vehicle_value,
    driver_age,
    vehicle_age,
    driving_experience,
    total_claims,
    total_exposure
  )

cor_matrix <- cor(
  numeric_vars,
  use = "complete.obs"
)

corrplot(
  cor_matrix,
  method = "color",
  type = "lower",
  addCoef.col = "black",
  tl.col = "black",
  tl.cex = 0.9,
  number.cex = 0.7,
  col = colorRampPalette(
    c("steelblue", "white", "tomato")
  )(200)
)
```

## The code

```{r}
#| eval: false

numeric_vars <- insurance_final %>%
  select(
    total_premium,
    total_incurred,
    net_profit,
    loss_ratio,
    claim_frequency,
    claim_severity,
    vehicle_value,
    driver_age,
    vehicle_age,
    driving_experience,
    total_claims,
    total_exposure
  )

cor_matrix <- cor(
  numeric_vars,
  use = "complete.obs"
)

corrplot(
  cor_matrix,
  method = "color",
  type = "lower",
  addCoef.col = "black",
  tl.col = "black",
  tl.cex = 0.9,
  number.cex = 0.7,
  col = colorRampPalette(
    c("steelblue", "white", "tomato")
  )(200)
)
```
:::

::: callout-note
Correlation analysis was conducted on selected variables associated with portfolio profitability and volatility.

Driver age and driving experience demonstrate relatively weak direct relationships with profitability-related variables, with absolute correlation values ranging only between 0.01 and 0.02. This suggests that customer demographic characteristics alone may have limited explanatory power in predicting underwriting performance.

Similarly, vehicle value and vehicle age also exhibit weak correlations with profitability metrics, indicating that vehicle characteristics do not independently drive portfolio profitability to a substantial extent.

Among the profitability-related variables, claim frequency demonstrates the strongest associations with driver and vehicle-related attributes, although the relationships remain relatively modest at approximately 0.02 in absolute value.
:::

## 5 EDA on Profitability Driver

This section applies appropriate bivariate visual analytics techniques to further investigate the relationship between variables and portfolio profitability. Further statistical test is applied to validate contributing factors.

## 5.1 Net Profit by Driver Age Group

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(
    x = net_profit,
    y = age_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = age_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(xlim = c(-5000, 3000)) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Net Profit Distribution Across Driver Age Groups",
    x = "Net Profit",
    y = "Driver Age Group"
  ) +
  theme_minimal(base_size = 14)

kruskal.test(
  net_profit ~ age_group,
  data = insurance_final
)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(
    x = net_profit,
    y = age_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = age_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(xlim = c(-5000, 3000)) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Net Profit Distribution Across Driver Age Groups",
    x = "Net Profit",
    y = "Driver Age Group"
  ) +
  theme_minimal(base_size = 14)

kruskal.test(
  net_profit ~ age_group,
  data = insurance_final
)
```
:::

::: callout-note
The net profit distribution across different driver age groups is largely similar according to the ridgeline and boxplot.

Drivers between 18 to 30 years old drivers with 0 to 5 years of experience appear to exhibit a narrower negative-profit tail as compared to the rest.

Since the distribution of age groups is not normal, a Kruskal-Wallis test was performed to evaluate whether net profit differs significantly across the groups at 5% significance level.

The test results shows that p value is \< 0.05, rejects the null hypothesis and suggests that net profit does differ significantly across driver age groups.
:::

## 5.2 Net Profit by Driving Experience Groups

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(
    x = net_profit,
    y = experience_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = experience_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(xlim = c(-5000, 3000)) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Net Profit Distribution Across Driving Experience Groups",
    x = "Net Profit",
    y = "Driving Experience Group"
  ) +
  theme_minimal(base_size = 14)

kruskal.test(
  net_profit ~ experience_group,
  data = insurance_final
)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(
    x = net_profit,
    y = experience_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = experience_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(xlim = c(-5000, 3000)) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Net Profit Distribution Across Driving Experience Groups",
    x = "Net Profit",
    y = "Driving Experience Group"
  ) +
  theme_minimal(base_size = 14)

kruskal.test(
  net_profit ~ experience_group,
  data = insurance_final
)
```
:::

::: callout-note
The net profit distribution across different driving experience groups is largely similar according to the ridgeline and boxplot.

Drivers with 0 to 5 years of experience appear to exhibit a narrower negative-profit tail as compared to the rest.

Since the distribution of driving experience groups is not normal, a Kruskal-Wallis test was performed to evaluate whether net profit differs significantly across the groups at 5% significance level.

The test results shows that p value is \< 0.05, rejects the null hypothesis and suggests that net profit does differ significantly across driver experience groups.
:::

## 5.3 Net Profit by Policy Types

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(
    x = net_profit,
    y = policy_type
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = policy_type),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(xlim = c(-5000, 3000)) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Net Profit Distribution Across Policy Types",
    x = "Net Profit",
    y = "Policy Type"
  ) +
  theme_minimal(base_size = 14)

kruskal.test(
  net_profit ~ policy_type,
  data = insurance_final
)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(
    x = net_profit,
    y = policy_type
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = policy_type),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(xlim = c(-5000, 3000)) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Net Profit Distribution Across Policy Types",
    x = "Net Profit",
    y = "Policy Type"
  ) +
  theme_minimal(base_size = 14)

kruskal.test(
  net_profit ~ policy_type,
  data = insurance_final
)
```
:::

::: callout-note
The net profit distribution across different policy types are varied according to the ridgeline and boxplot.

Notably, basic third part liability coverage exhibits more concentrated profitability distributions while the rest have heavier negative and positive profit tails.

Since the distribution of policy types is not normal, a Kruskal-Wallis test was performed to evaluate whether net profit differs significantly across the groups at 5% significance level.

The test results shows that p value is \< 0.05, rejects the null hypothesis and suggests that net profit does differ significantly across policy types.
:::

## 6 Volatility & Loss Driver Analysis

This section applies appropriate bivariate visual analytics techniques to further investigate the relationship between variables and portfolio volatility. Further statistical test is applied to validate contributing factors.

## 6.1 Claim Frequency by Vehicle Age Groups

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(
    x = claim_frequency,
    y = vehicle_age_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = vehicle_age_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(xlim = c(0, 5)) +
  labs(
    title = "Claim Frequency Distribution Across Vehicle Age Groups",
    x = "Claim Frequency",
    y = "Vehicle Age Group"
  ) +
  theme_minimal(base_size = 14)

kruskal.test(
  claim_frequency ~ vehicle_age_group,
  data = insurance_final
)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(
    x = claim_frequency,
    y = vehicle_age_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = vehicle_age_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(xlim = c(0, 5)) +
  labs(
    title = "Claim Frequency Distribution Across Vehicle Age Groups",
    x = "Claim Frequency",
    y = "Vehicle Age Group"
  ) +
  theme_minimal(base_size = 14)

kruskal.test(
  claim_frequency ~ vehicle_age_group,
  data = insurance_final
)
```
:::

::: callout-note
Older vehicles exhibit wider and more volatile claim frequency distributions, suggesting increased accident occurrence risk. Newer vehicles demonstrate relatively concentrated claim frequency behaviour, indicating more stable risk exposure.

Since the distribution of vehicle age groups is not normal, a Kruskal-Wallis test was performed to evaluate whether claim frequency differs significantly across the groups at 5% significance level.

The test results shows that p value is \< 0.05, rejects the null hypothesis and suggests that claim frequency does differ significantly across vehicle age groups.
:::

## 6.2 Claim Frequency by Driving Experience Groups

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(
    x = claim_frequency,
    y = experience_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = experience_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(
    xlim = c(0, 5)
  ) +
  labs(
    title = "Claim Frequency Distribution Across Driving Experience Groups",
    x = "Claim Frequency",
    y = "Driving Experience Group"
  ) +
  theme_minimal(base_size = 14)

kruskal.test(
  claim_frequency ~ experience_group,
  data = insurance_final
)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(
    x = claim_frequency,
    y = experience_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = experience_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(
    xlim = c(0, 5)
  ) +
  labs(
    title = "Claim Frequency Distribution Across Driving Experience Groups",
    x = "Claim Frequency",
    y = "Driving Experience Group"
  ) +
  theme_minimal(base_size = 14)

kruskal.test(
  claim_frequency ~ experience_group,
  data = insurance_final
)
```
:::

::: callout-note
Drivers with higher driving experience appear to exhibit more outliers in frequency boxplot, suggesting increased accident occurrence volatility and greater exposure to repeated claim events.

Since the distribution of driving experience groups is not normal, a Kruskal-Wallis test was performed to evaluate whether claim frequency differs significantly across the groups at 5% significance level.

The test results shows that p value is \< 0.05, rejects the null hypothesis and suggests that claim frequency does differ significantly across driving experience groups.
:::

## 6.3 Claim Severity by Vehicle Age Groups

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(
    x = claim_severity,
    y = vehicle_age_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = vehicle_age_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(
    xlim = c(0, 20000)
  ) +
  scale_x_continuous(
    labels = scales::comma
  ) +
  labs(
    title = "Claim Severity Distribution Across Vehicle Age Groups",
    x = "Claim Severity",
    y = "Vehicle Age Group"
  ) +
  theme_minimal(base_size = 14)

kruskal.test(
  claim_severity ~ vehicle_age_group,
  data = insurance_final
)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(
    x = claim_severity,
    y = vehicle_age_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = vehicle_age_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(
    xlim = c(0, 20000)
  ) +
  scale_x_continuous(
    labels = scales::comma
  ) +
  labs(
    title = "Claim Severity Distribution Across Vehicle Age Groups",
    x = "Claim Severity",
    y = "Vehicle Age Group"
  ) +
  theme_minimal(base_size = 14)

kruskal.test(
  claim_severity ~ vehicle_age_group,
  data = insurance_final
)
```
:::

::: callout-note
The claim severity distribution across different vehicle age groups demonstrates visible variability according to the ridgeline and boxplot.

Older vehicles appear to exhibit heavier positive claim severity tails, suggesting greater exposure to severe financial losses and elevated underwriting volatility.

Since the distribution of vehicle age groups is not normal, a Kruskal-Wallis test was performed to evaluate whether claim severity differs significantly across the groups at 5% significance level.

The test results shows that p value is \< 0.05, rejects the null hypothesis and suggests that claim severity does differ significantly across vehicle age groups.
:::

## 6.4 Claim Severity by Driving Experience Groups

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  insurance_final,
  aes(
    x = claim_severity,
    y = experience_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = experience_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(
    xlim = c(0, 20000)
  ) +
  scale_x_continuous(
    labels = scales::comma
  ) +
  labs(
    title = "Claim Severity Distribution Across Driving Experience Groups",
    x = "Claim Severity",
    y = "Driving Experience Group"
  ) +
  theme_minimal(base_size = 14)

kruskal.test(
  claim_severity ~ experience_group,
  data = insurance_final
)
```

## The code

```{r}
#| eval: false

ggplot(
  insurance_final,
  aes(
    x = claim_severity,
    y = experience_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = experience_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(
    xlim = c(0, 20000)
  ) +
  scale_x_continuous(
    labels = scales::comma
  ) +
  labs(
    title = "Claim Severity Distribution Across Driving Experience Groups",
    x = "Claim Severity",
    y = "Driving Experience Group"
  ) +
  theme_minimal(base_size = 14)

kruskal.test(
  claim_severity ~ experience_group,
  data = insurance_final
)
```
:::

::: callout-note
The claim severity distribution across different driving experience groups is relatively similar according to the ridgeline and boxplot.

Drivers with higher driving experience appear to have more outliers in severity boxplot, suggesting greater exposure to severe claim events and increased underwriting volatility.

Since the distribution of driving experience groups is not normal, a Kruskal-Wallis test was performed to evaluate whether claim severity differs significantly across the groups at 5% significance level.

The test results shows that p value is \< 0.05, rejects the null hypothesis and suggests that claim severity does differ significantly across driving experience groups.
:::

## 7 Temporal Trend Analysis

### 7.1 Profitability Trend

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

yearly_summary <- insurance_final %>%
  group_by(year) %>%
  summarise(
    avg_profit = mean(net_profit, na.rm = TRUE)
  )

ggplot(
  yearly_summary,
  aes(
    x = year,
    y = avg_profit,
    group = 1
  )
) +
  geom_line(
    color = "steelblue",
    linewidth = 1.2
  ) +
  geom_point(
    color = "steelblue",
    size = 3
  ) +
  labs(
    title = "Average Net Profit Across Years",
    x = "Year",
    y = "Average Net Profit"
  ) +
  theme_minimal(base_size = 14)
```

## The code

```{r}
#| eval: false

yearly_summary <- insurance_final %>%
  group_by(year) %>%
  summarise(
    avg_profit = mean(net_profit, na.rm = TRUE)
  )

ggplot(
  yearly_summary,
  aes(
    x = year,
    y = avg_profit,
    group = 1
  )
) +
  geom_line(
    color = "steelblue",
    linewidth = 1.2
  ) +
  geom_point(
    color = "steelblue",
    size = 3
  ) +
  labs(
    title = "Average Net Profit Across Years",
    x = "Year",
    y = "Average Net Profit"
  ) +
  theme_minimal(base_size = 14)
```
:::

::: callout-note
The average net profit across the years increased from 2022 to 2023, but decreased sharply in 2024. This suggests that underwriting performance initially improved but later deteriorated significantly over time.
:::

### 7.2 Claim Trend

This section examines the temporal movement of portfolio profitability and claim behaviour across the observation years.

::: panel-tabset
## The plot

```{r}
#| echo: false
#| message: false
#| warning: false

claim_trend <- insurance_final %>%
  group_by(year) %>%
  summarise(
    avg_incurred = mean(total_incurred, na.rm = TRUE),
    avg_claim_frequency = mean(claim_frequency, na.rm = TRUE)
  )

ggplot(
  claim_trend,
  aes(x = year, group = 1)
) +
  geom_line(
    aes(y = avg_incurred, color = "Average Incurred Loss")
  ) +
  geom_line(
    aes(y = avg_claim_frequency * 1000,
        color = "Claim Frequency (Scaled)")
  ) +
  labs(
    title = "Claim Trend Across Years",
    x = "Year",
    y = "Trend Value",
    color = "Metric"
  ) +
  theme_minimal(base_size = 14)
```

## The code

```{r}
#| eval: false

claim_trend <- insurance_final %>%
  group_by(year) %>%
  summarise(
    avg_incurred = mean(total_incurred, na.rm = TRUE),
    avg_claim_frequency = mean(claim_frequency, na.rm = TRUE)
  )

ggplot(
  claim_trend,
  aes(x = year, group = 1)
) +
  geom_line(
    aes(y = avg_incurred, color = "Average Incurred Loss")
  ) +
  geom_line(
    aes(y = avg_claim_frequency * 1000,
        color = "Claim Frequency (Scaled)")
  ) +
  labs(
    title = "Claim Trend Across Years",
    x = "Year",
    y = "Trend Value",
    color = "Metric"
  ) +
  theme_minimal(base_size = 14)

```
:::

::: callout-note
The claim trend analysis suggests that claim frequency increased steadily from 2022 to 2024, indicating a gradual rise in accident occurrence rates and overall portfolio risk exposure over time.

At the same time, average incurred losses also increased progressively across the years, with a noticeably steeper increase between 2023 and 2024. This suggests that not only are claims occurring more frequently, but the financial severity of claims may also be worsening.
:::

# 8. Conclusion

In conclusion, the analysis revealed that the portfolio is dominated by low-claim and low-loss policies, while a relatively small proportion of policies contributes disproportionately large financial losses.

Profitability analysis showed that policy type demonstrates clearer differences in portfolio performance as compared to individual demographic or vehicle characteristics alone. Although driver age, driving experience, vehicle age, and vehicle value exhibit weak linear correlations with profitability metrics, statistical testing confirmed that net profit still differs significantly across the assessed customer and vehicle variables.

Volatility analysis showed that older vehicles and more experienced drivers are associated with wider claim frequency and claim severity distributions.

Temporal trend analysis revealed deteriorating portfolio performance over time. Claim frequency and incurred losses increased steadily from 2022 to 2024, while average profitability declined sharply in 2024.

Overall, the visual analytics techniques have uncovered hidden portfolio risk patterns and profitability drivers.
