---
title: "Take-home Exercise 1"
subtitle: "Visual Analytics for Motor Insurance Portfolio Discovery"
author: "Michelle Chen"
date: "23 May 2026"
format:
  revealjs:
    theme: simple
    slide-number: true
    smaller: true
    scrollable: true
    css: style.css
    preview-links: auto
    footer: "Visual Analytics for Motor Insurance Portfolio Discovery"
execute:
  echo: false
  warning: false
  message: false
editor: visual
---

```{r}
#| include: false
pacman::p_load(tidyverse, ggplot2, scales, ggridges, corrplot, patchwork)

# Use ; to identify separate columns in the csv file
insurance_data <- read_delim(
  "data/Dataset of motor insurance portfolio.csv", 
  delim = ";",
  escape_double = FALSE,
  trim_ws = TRUE
)

# Calculate missing values for all columns
missing_counts <- colSums(is.na(insurance_data))
# Display columns that have at least 1 missing value
missing_counts[missing_counts > 0]

# Drop any row that contains an NA in these specific columns
insurance_clean <- insurance_data %>%
  drop_na(vehicle_age, age_driving_licence, fuel_type, vehicle_value)
```

```{r setup}
#| include: false

# Load packages
pacman::p_load(
  tidyverse,
  scales,
  ggridges,
  corrplot,
  patchwork
)

# Load dataset
insurance_data <- read_delim(
  "data/Dataset of motor insurance portfolio.csv",
  delim = ";",
  escape_double = FALSE,
  trim_ws = TRUE,
  show_col_types = FALSE
)

# Drop missing values
insurance_clean <- insurance_data %>%
  drop_na(vehicle_age, age_driving_licence, fuel_type, vehicle_value)

# Transform datatype
insurance_clean <- insurance_clean %>%
  mutate(
# Convert unique ID into text to prevent accidental math operations
    insured_id = as.character(insured_id), 
# Convert the tracking year into categorical factors
    year = as.factor(year),
# Convert all text character columns into categorical factors automatically
    across(
      where(is.character) & -insured_id, as.factor
  )
)

# Feature engineering
insurance_final <- insurance_clean %>%
  mutate(
    # 1. Net Profit
    net_profit = total_premium - total_incurred,
    
    # 2. Loss Ratio
    loss_ratio = if_else(
      total_premium > 0, total_incurred / total_premium, NA_real_
    ),
    
    # 3. Claim Frequency
    claim_frequency = if_else(
      total_exposure > 0.0, total_claims / total_exposure, NA_real_
    ),
    
    # 4. Claim Severity
    claim_severity = if_else(
      total_claims > 0.0, total_incurred / total_claims, NA_real_
    ),
    
    # 5. Claim indicator 
    has_claim = factor(
      if_else(total_claims > 0.0, "Yes", "No")
    ),
    
    # 6. Driving experience
    driving_experience = driver_age - age_driving_licence,
    experience_group = cut(
      driving_experience,
      breaks = c(0, 5, 10, 20, 30, Inf),
      labels = c("0-5", "6-10", "11-20", "21-30", "30+"),
      include.lowest = TRUE 
    ),
    
    # 7. Vehicle age group 
    vehicle_age_group = cut(
      vehicle_age,
      breaks = c(0, 5, 10, 15, 20, Inf),
      labels = c("0-5", "6-10", "11-15", "16-20", "20+"),
      include.lowest = TRUE  #
    ),
    
    # 8. Driver age group
    age_group = cut(
      driver_age,
      breaks = c(18, 30, 40, 50, 60, Inf),
      labels = c("18-30", "31-40", "41-50", "51-60", "60+"),
      include.lowest = TRUE  
    )
  )
```

## Contents

1.  Financial Variable - Total Premium

2.  Financial Variable - Incurred Losses, Claim Severity

3.  Customer Demographics - Driver Age and Experience

4.  Customer Demographics - Driver Bonus Score

5.  Vehicle Characteristics - Value and Age

6.  Portfolio Composition - Policy Type, Claim Occurrence

7.  Correlation Matrix

8.  Net Profit by Driver Age and Experience

9.  Net Profit by Policy Type

10. Claim Frequency and Severity by Vehicle Age

11. Claim Frequency and Severity by Driving Experience

12. Profitability and Claim Trend

13. Conclusion

## Financial Variable - Total Premium

```{r}
#| label: totalpremium
#| fig-width: 13
#| fig-height: 5
#| echo: false
#| warning: false
#| message: false


ggplot(
  insurance_final,
  aes(x = total_premium)
) +
  geom_histogram(
    bins = 40,
    fill = "steelblue",
    color = "white"
  ) +
  coord_cartesian(xlim = c(0, 1500)) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Distribution of Total Premium",
    x = "Total Premium",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 14)
```

::: aside
- Right-skewed distribution - most policyholders contribute low premium amounts, \< \$500/year.

- A smaller group contributes larger premiums, up to around \$1 250/year.

- The customer segments have varying levels of risk exposure and coverage requirements.
:::

## Financial Variable - Incurred Losses, Claim Severity

```{r}
#| label: lossesxclaimseverity
#| fig-width: 13
#| fig-height: 5
#| echo: false
#| warning: false
#| message: false

p1 <- ggplot(
  insurance_final,
  aes(x = log1p(total_incurred))
) +
  geom_histogram(
    bins = 50,
    fill = "steelblue",
    color = "white"
  ) +
  labs(
    title = "Distribution of Total Incurred Losses on Log Scale",
    x = "Log(1 + Total Incurred Losses)",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 14)

p2 <- ggplot(
  insurance_final,
  aes(x = claim_severity)
) +
  geom_histogram(
    bins = 50,
    fill = "steelblue",
    color = "white"
  ) +
  coord_cartesian(
    xlim = c(0, 20000)
  ) +
  scale_x_continuous(
    labels = scales::comma
  ) +
  labs(
    title = "Distribution of Claim Severity",
    x = "Claim Severity",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 14)

patchwork::wrap_plots(p1, p2, ncol = 2)
```

::: aside
- Policies with little or no financial losses dominates the portfolio.

- Claim severity contains extreme outliers - a small number of claims generate disproportionately high financial losses.
:::

## Customer Demographics - Driver Age and Experience

```{r}
#| label: driveragexexperience
#| fig-width: 13
#| fig-height: 5
#| echo: false
#| warning: false
#| message: false

p1 <- ggplot(
  insurance_final,
  aes(x = age_group)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Driver Age Groups",
    x = "Driver Age Group",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)

p2 <- ggplot(
  insurance_final,
  aes(x = experience_group)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Driving Experience Groups",
    x = "Driving Experience Group",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)

patchwork::wrap_plots(p1, p2, ncol = 2)
```

::: aside
- Most policyholders are in the 41 to 50 years old age group while a very small portion of them are 30 years old and below.

- Experienced drivers dominate the portfolio - most have more than 10 years of driving experience.
:::

## Customer Demographics - Driver Bonus Score

```{r}
#| label: driverbonusscore
#| fig-width: 13
#| fig-height: 5
#| echo: false
#| warning: false
#| message: false

ggplot(
  insurance_final,
  aes(x = bonus_score)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Bonus Score",
    x = "Bonus Score",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)
```

::: aside
- Most of the policyholders have a positive bonus score.

- This reflects that they have favourable claims history and are generally less risky policyholders.
:::

## Vehicle Characteristics - Value and Age

```{r}
#| label: vehiclevaluexage
#| fig-width: 13
#| fig-height: 5
#| echo: false
#| warning: false
#| message: false

p1 <- ggplot(
  insurance_final,
  aes(x = vehicle_value)
) +
  geom_histogram(
    bins = 40,
    fill = "steelblue",
    color = "white"
  ) +
  coord_cartesian(xlim = c(0, 50000)) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Distribution of Vehicle Value",
    x = "Vehicle Value",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 14)

p2 <- ggplot(
  insurance_final,
  aes(x = vehicle_age_group)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Vehicle Age Groups",
    x = "Vehicle Age Group",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)

patchwork::wrap_plots(p1, p2, ncol = 2)
```

::: aside
- The vehicle values are right skewed, with most of them ranging between \$15 000 to \$35 000.

- Most insured vehicles are of moderate market values while a smaller segment containing substantially higher-valued vehicles.

- Most insured vehicles are 20 years old and above, which may suggests that they have higher maintenance and operational risks. A statistical test is performed to validate this hypothesis in later slides.
:::

## Portfolio Composition - Policy Type, Claim Occurrence

```{r}
#| label: policytypexclaim
#| fig-width: 13
#| fig-height: 5
#| echo: false
#| warning: false
#| message: false

p1 <- ggplot(
  insurance_final,
  aes(x = policy_type)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Policy Types",
    x = "Policy Type",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)

p2 <- ggplot(
  insurance_final,
  aes(x = has_claim)
) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Claim Occurrence",
    x = "Claim Indicator",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)

patchwork::wrap_plots(p1, p2, ncol = 2)
```

::: aside
- Third-party liability combined with two or more additional coverage is the most common insurance purchased.

- The portfolio is dominated by non-claim policies - claims occur infrequently.
:::

## Correlation Matrix

```{r}
#| label: correlationmatrix
#| fig-width: 13
#| fig-height: 5
#| echo: false
#| warning: false
#| message: false

numeric_vars <- insurance_final %>%
  select(
    total_premium,
    total_incurred,
    net_profit,
    loss_ratio,
    claim_frequency,
    claim_severity,
    vehicle_value,
    driver_age,
    vehicle_age,
    driving_experience,
    total_claims,
    total_exposure
  )

cor_matrix <- cor(
  numeric_vars,
  use = "complete.obs"
)

corrplot(
  cor_matrix,
  method = "color",
  type = "lower",
  addCoef.col = "black",
  tl.col = "black",
  tl.cex = 0.9,
  number.cex = 0.7,
  col = colorRampPalette(
    c("steelblue", "white", "tomato")
  )(200)
)
```

::: aside
- Correlation matrix analysis is done on numerical variables.

- Unexpectedly, driver age and driving experience demonstrate weak linear relationships with profitability-related variables (absolute values between 0.01 and 0.02).

- Vehicle value and vehicle age also exhibit weak correlations with profitability-related variables.

- This suggest that any variable alone does not have strong correlation with profitability and volatility measures.
:::

## Net Profit by Driver Age and Experience

```{r}
#| label: netprofitxdriver
#| fig-width: 13
#| fig-height: 5
#| echo: false
#| warning: false
#| message: false

p1 <- ggplot(
  insurance_final,
  aes(
    x = net_profit,
    y = age_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = age_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(xlim = c(-5000, 3000)) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Net Profit Distribution Across Driver Age Groups",
    x = "Net Profit",
    y = "Driver Age Group"
  ) +
  theme_minimal(base_size = 14)

p2 <- ggplot(
  insurance_final,
  aes(
    x = net_profit,
    y = experience_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = experience_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(xlim = c(-5000, 3000)) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Net Profit Distribution Across Driving Experience Groups",
    x = "Net Profit",
    y = "Driving Experience Group"
  ) +
  theme_minimal(base_size = 14)

patchwork::wrap_plots(p1, p2, ncol = 2)
```

::: aside
- The net profit distribution across different driver age groups and years of experience appears to be largely similar.

- Drivers between 18 to 30 years old drivers with 0 to 5 years of experience exhibits a narrower negative-profit tail as compared to the rest.

- A statistical test validates that **net** **profit differs significantly** across age groups and driving experience groups.
:::

## Net Profit by Policy Type

```{r}
#| label: policytype
#| fig-width: 13
#| fig-height: 5
#| echo: false
#| warning: false
#| message: false

ggplot(
  insurance_final,
  aes(
    x = net_profit,
    y = policy_type
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = policy_type),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(xlim = c(-5000, 3000)) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Net Profit Distribution Across Policy Types",
    x = "Net Profit",
    y = "Policy Type"
  ) +
  theme_minimal(base_size = 14)
```

::: aside
- Basic third party liability coverage exhibits more concentrated profitability distributions than the rest.

- A statistical test validates that **net profit differs significantly** across policy types.
:::

## Claim Frequency and Severity by Vehicle Age

```{r}
#| label: claimxvehicleage
#| fig-width: 13
#| fig-height: 5
#| echo: false
#| warning: false
#| message: false

p1 <- ggplot(
  insurance_final,
  aes(
    x = claim_frequency,
    y = vehicle_age_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = vehicle_age_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(xlim = c(0, 5)) +
  labs(
    title = "Claim Frequency by Vehicle Age Groups",
    x = "Claim Frequency",
    y = "Vehicle Age Group"
  ) +
  theme_minimal(base_size = 14)

p2 <- ggplot(
  insurance_final,
  aes(
    x = claim_severity,
    y = vehicle_age_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = vehicle_age_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(
    xlim = c(0, 20000)
  ) +
  scale_x_continuous(
    labels = scales::comma
  ) +
  labs(
    title = "Claim Severity by Vehicle Age Groups",
    x = "Claim Severity",
    y = "Vehicle Age Group"
  ) +
  theme_minimal(base_size = 14)

patchwork::wrap_plots(p1, p2, ncol = 2)
```

::: aside
- Older vehicles exhibit wider and more volatile claim frequency distributions, suggesting increased accident occurrence risk.

- Older vehicles also exhibit heavier positive claim severity tails, suggesting greater exposure to severe financial losses.

- A statistical test validates that **claim frequency AND severity differs significantly** across vehicle age groups.
:::

## Claim Frequency and Severity by Driving Experience

```{r}
#| label: claimxdrivingexperience
#| fig-width: 13
#| fig-height: 4
#| echo: false
#| warning: false
#| message: false

p1 <- ggplot(
  insurance_final,
  aes(
    x = claim_frequency,
    y = experience_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = experience_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(
    xlim = c(0, 5)
  ) +
  labs(
    title = "Claim Frequency by Driving Experience Groups",
    x = "Claim Frequency",
    y = "Driving Experience Group"
  ) +
  theme_minimal(base_size = 14)

p2 <- ggplot(
  insurance_final,
  aes(
    x = claim_severity,
    y = experience_group
  )
) +
  ggridges::geom_density_ridges(
    fill = "steelblue",
    alpha = 0.7,
    scale = 1.2,
    color = NA
  ) +
  geom_boxplot(
    aes(group = experience_group),
    width = 0.12,
    fill = "gray",
    color = "black",
    outlier.alpha = 0.15
  ) +
  coord_cartesian(
    xlim = c(0, 20000)
  ) +
  scale_x_continuous(
    labels = scales::comma
  ) +
  labs(
    title = "Claim Severity by Driving Experience Groups",
    x = "Claim Severity",
    y = "Driving Experience Group"
  ) +
  theme_minimal(base_size = 14)

patchwork::wrap_plots(p1, p2, ncol = 2)
```

::: aside
- Drivers with higher driving experience have more outliers in frequency boxplot - suggests increased accident occurrence volatility.

- Drivers with higher driving experience also have more outliers in claim severity tails - suggests greater exposure to severe claim events.

- A statistical test validates that **claim frequency AND severity differs significantly** across driving experience groups.
:::

## Profitability and Claim Trend 

```{r}
#| label: profitxclaimtrend
#| fig-width: 13
#| fig-height: 5
#| echo: false
#| warning: false
#| message: false

yearly_summary <- insurance_final %>%
  group_by(year) %>%
  summarise(
    avg_profit = mean(net_profit, na.rm = TRUE)
  )

p1 <- ggplot(
  yearly_summary,
  aes(
    x = year,
    y = avg_profit,
    group = 1
  )
) +
  geom_line(
    color = "steelblue",
    linewidth = 1.2
  ) +
  geom_point(
    color = "steelblue",
    size = 3
  ) +
  labs(
    title = "Average Net Profit Across Years",
    x = "Year",
    y = "Average Net Profit"
  ) +
  theme_minimal(base_size = 14)

claim_trend <- insurance_final %>%
  group_by(year) %>%
  summarise(
    avg_incurred = mean(total_incurred, na.rm = TRUE),
    avg_claim_frequency = mean(claim_frequency, na.rm = TRUE)
  )

p2 <- ggplot(
  claim_trend,
  aes(x = year, group = 1)
) +
  geom_line(
    aes(y = avg_incurred, color = "Average Incurred Loss")
  ) +
  geom_line(
    aes(y = avg_claim_frequency * 1000,
        color = "Claim Frequency (Scaled)")
  ) +
  labs(
    title = "Claim Trend Across Years",
    x = "Year",
    y = "Trend Value",
    color = "Metric"
  ) +
  theme_minimal(base_size = 14)

patchwork::wrap_plots(p1, p2, ncol = 2)
```

::: aside
- Claim frequency increased steadily from 2022 to 2024, indicating a gradual rise in accident occurrence rates and overall portfolio risk exposure over time.
- Average incurred losses also increased progressively with a steeper increase between 2023 and 2024.
- This suggests that not only are claims occurring more frequently, but the financial severity of claims may also be worsening.
:::

## Conclusion

1.  Portfolio is dominated by low-claim and low-loss policies.
2.  Policy type demonstrates clearer differences in portfolio performance as compared to individual demographic or vehicle characteristics alone.
3.  Portfolio net profit differs significantly across driver age, driving experience, vehicle age, and vehicle value.
4.  Volatility analysis showed that older vehicles and more experienced drivers are associated with wider claim frequency and claim severity distributions.
5.  There is a deteriorating portfolio performance over time:
    - claim frequency and incurred losses increased steadily from 2022 to 2024

    - average profitability declined sharply in 2024.
