---
title: "Take Home Ex2: VAST Challenge 2026 Mini-Challenge 1"
author: "Michelle Chen"
date: "6 June 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: false
format:
  html:
    code-fold: true
---

# 1. Overview

## 1.1 Business question

This exercise is focused on solving [Mini Challenge 1](https://vast-challenge.github.io/2026/MC1.html) of VAST Challenge 2026.

It digs into TenantThread's internal communications from the two weeks leading up to the embargo breach and investigates whether the team deliberately leaked the CivicLoom merger - **Project HarborCrest**, or whether it happened because the communication system broke down under pressure.

The analysis seeks to answer the following questions:

1.  What were the key events and relationships that led up to the inappropriate release?
2.  What are the typical behaviour of systems/agents/people involved and how does the behaviour that led to the release compare to the prior behaviors?
3.  Were there leading indicators that such a release was possible and if yes, why didn't the prior occasions result in noticeable action?

## 1.2 Methodology

This investigation adopts a visual analytics approach to reconstruct the sequence of events leading up to the embargo breach. The analysis combines temporal, network, behavioural, and text analytics techniques to examine:

1.  understanding how information propagated through the multi-agent communication environment
2.  identifying abnormal behaviours
3.  assessing whether the breach resulted from deliberate actions or systemic failures.

| Analytical Method | Visual Analytics Technique | Investigation Objective |
|:---|:---|:---|
| Temporal Analysis | Timeline visualisation and communication volume analysis | Identify key events, communication surges, and activity patterns leading up to the breach |
| Network Analysis | Agent-channel networks and peer-to-peer interaction networks | Examine information flow pathways and identify influential communication channels |
| Behavioural Analysis | Agent activity comparison and channel usage comparison | Detect shifts in communication behaviour before and during the breach |
| Text Analytics | Keyword tracking, sensitive topic analysis, and TF-IDF analysis | Identify leading indicators, trace the spread of sensitive information, and uncover agent-specific discussion themes |

# 2. Set Up

## 2.1 Load packages

Several packages are required to support the analysis in this exercise. The step below loads the necessarily ones into Rstudio.

```{r}
pacman::p_load(
  jsonlite,     # Parse JSON schemas
  tidyverse,    # For data wrangling 
  tidytext,     # For TD-IDF
  lubridate,    # Timeline manipulation
  tidygraph,    # For relational network graphs
  visNetwork,   # Interactive dynamic graph configurations
  plotly,       # Advanced interactive charting layers
)
```

## 2.2 Data import

The dataset is imported directly from [Mini Challenge 1](https://vast-challenge.github.io/2026/MC1.html) of Vast Challenge 2016.

```{r}
mc1_data <- fromJSON("data/MC1_final_00.json", simplifyVector = FALSE)
glimpse(mc1_data)
```

::: callout-note
There are a total of 23 crisis rounds, each capturing the event context, communication records, and participating agents. Since the communications and participants fields are stored as nested lists, further data wrangling is required to extract these details into a table format for deeper analysis.
:::

# 3. Data Wrangling

## 3.1 Round-Level Summary Table

The step below transforms raw round-level JSON data into a flattened round-level summary table to facilitate analysis and trends thorught the investigation period. This table captures the key contextual information associated with each crisis round.

```{r}
rounds_tbl <- map_df(mc1_data$rounds, function(r) {
  tibble(
    round_id = r$round_id %||% NA_integer_,
    round_time = ymd_hms(r$hour),
    event_headline = r$environment_context$event_headline,
    event_narrative = r$environment_context$event_narrative,
    social_state = r$environment_context$social_state,
    stock_price = as.numeric(str_remove(r$environment_context$market_snapshot$stock_price, "\\$")),
    n_communications = length(r$communications)
  )
}) %>%
  mutate(round_id = row_number())

glimpse(rounds_tbl)
```

## 3.2 Communication-Level Table

A communication-level table is also constructed by extracting all communication records from the simulation logs. It captures the content, timing, participants, communication channels, and internal delibrations associated with each message. This table will be used to analyse behavioural, network, and keyword investigations.

```{r}
communications_tbl <- map_df(mc1_data$rounds, function(r) {
  map_df(r$communications, function(c) {
    tibble(
      round_id = r$round_id %||% NA_integer_,
      round_time = ymd_hms(r$hour),
      timestamp = ymd_hms(c$timestamp),
      message_id = c$message_id %||% NA_character_,
      agent_id = c$agent_id %||% NA_character_,
      agent_role = c$agent_role %||% NA_character_,
      agent_label = c$agent_label %||% NA_character_,
      channel = c$channel %||% NA_character_,
      message_type = c$message_type %||% NA_character_,
      content = c$content %||% NA_character_,
      recipients = list(c$recipients %||% character(0)),
      deliberation = c$internal_state$deliberating %||% NA_character_,
      rationalization = c$internal_state$rationalizing %||% NA_character_
    )
  })
}) %>%
  mutate(
    round_id = if_else(is.na(round_id), dense_rank(round_time), round_id),
    round_date = as.Date(timestamp),
    content_lower = str_to_lower(content),
    internal_text = str_c(deliberation, rationalization, sep = " ")
  )

glimpse(communications_tbl)
```

## 3.3 Network Nodes and Edges

This section transforms message records into a direct agent-channel network. This facilitates the analysis of how information moved through TenantThread’s communication system. Agents were treated as source nodes, communication channels as target nodes, and message counts as weighted edges.

```{r}
edges_tbl <- communications_tbl %>%
  filter(!is.na(channel), !is.na(agent_label)) %>%
  count(agent_label, channel, name = "weight") %>%
  rename(from = agent_label, to = channel)

nodes_tbl <- tibble(
  id = unique(c(edges_tbl$from, edges_tbl$to))
) %>%
  mutate(
    type = if_else(id %in% edges_tbl$from, "Agent", "Channel"),
    group = type,
    label = id
  )

comms_graph <- tbl_graph(
  nodes = nodes_tbl,
  edges = edges_tbl,
  directed = TRUE
) %>%
  activate(nodes) %>%
  mutate(
    degree_centrality = centrality_degree(mode = "all"),
    betweenness_stat = centrality_betweenness()
  )

glimpse(nodes_tbl)
glimpse(edges_tbl)
```

# 4 Key Events and Relationships Leading to the Embargo Breach

## 4.1 Identify Key Events

First, it is important to identify major organisational, governance, and reputational events that occurred before the embargo breach. Communication volume is used to highlight events that generated significant attention and response.

```{r}
key_events_tbl <- rounds_tbl %>%
  filter(as.Date(round_time) < as.Date("2046-06-05")) %>%
  select(
    round_time,
    event_headline,
    n_communications
  )

ggplot(
  key_events_tbl,
  aes(
    x = round_time,
    y = reorder(event_headline, round_time),
    size = n_communications
  )
) +
  geom_point() +
  labs(
    title = "Key Events Before Embargo Breach",
    x = "Timeline",
    y = "Event"
  ) +
  theme_minimal()
```

::: callout-note
Each round in the dataset was anchored by a dominant organisational event. Based on the events listed, external scrutiny progressively intensifies during the two weeks period. Notably, events such as the Crestview demo, NHPI report, usage guidelines delivery, the deployment of Judge, and the SaltWind Pieces generated significant communication spikes - 35 to 40. The system was operating under increasing pressure from both external events and heightened communication levels.
:::

## 4.2 Breach Day Communication Trend

Next, the analysis takes a closer look into the communication volume during breach day.

```{r}
breach_day_tbl <- communications_tbl %>%
  filter(round_date == as.Date("2046-06-05")) %>%
  mutate(
    hour = lubridate::floor_date(timestamp, "hour")
  ) %>%
  count(hour)

plot_ly(
  data = breach_day_tbl,
  x = ~hour,
  y = ~n,
  type = "scatter",
  mode = "lines+markers",
  hovertemplate =
    paste(
      "<b>Time:</b> %{x}",
      "<br><b>Messages:</b> %{y}",
      "<extra></extra>"
    )
) %>%
  layout(
    title = list(text = "Communication Activity on Breach Day"),
    xaxis = list(
      title = "Time",
      showgrid = TRUE
    ),
    yaxis = list(
      title = "Messages",
      showgrid = TRUE
    )
  )
```

::: callout-note
Communication activity remained consistently high throughout breach day. Notably, communication volume surged to its peak at 12pm on breach day, dropped sharply at 1pm before surging again from 2pm to 3pm, just 2 hours before the inappropriate release happened. This suggests that organisational attention became increasingly focused on merger-related matters approaching the release.
:::

## 4.3 Spread of Sensitive Information

In this step, merger-related and embargo-related discussions were analysed across monitored and unmonitored communication spaces to determine whether sensitive information remained contained.

```{r}
bypass_analysis_df <- communications_tbl %>%
  filter(
    str_detect(
      str_to_lower(content),
      "civicloom|harborcrest|merger|embargo"
    )
  ) %>%
  mutate(
    channel_type = case_when(
      channel %in% c("official_post", "comms_huddle") ~
        "Official Channels",
      TRUE ~
        "Unofficial Channels"
    )
  ) %>%
  count(round_id, channel_type) %>%
  left_join(
    rounds_tbl %>%
      select(round_id, round_time),
    by = "round_id"
  ) %>%
  filter(!is.na(round_time))

ggplot(
  bypass_analysis_df,
  aes(
    x = round_time,
    y = n,
    fill = channel_type
  )
) +
  geom_area(
    position = "stack",
    alpha = 0.8
  ) +
  scale_fill_manual(
    values = c(
      "Official Channels" = "#34495e",
      "Unofficial Channels" = "#e67e22"
    )
  ) +
  labs(
    title = "Spread of Sensitive Information",
    subtitle = "Merger and embargo discussions increasingly appeared outside official channels",
    x = "Timeline",
    y = "Sensitive Topic Mentions",
    fill = "Channel Type"
  ) +
  theme_minimal()
```

::: callout-note
Sensitive information related to the merger occurs in unofficial channels throughout the two weeks period. The orange areas shows that information was circulating through less-controlled spaces before the breach, and even during the breach day. This weakens the effectiveness of embargo controls and increases the likelihood of disclosure.
:::

## 4.4 Agent Participation Across Crisis Rounds

Agent participation was analysed to identify the actors most actively involved throughout the investigation period. It can further determine which agents played central roles in discussions and decision-making leading up to the embargo breach.

```{r}
agent_participation_tbl <- communications_tbl %>%
  distinct(round_id, agent_label) %>%
  count(agent_label, name = "rounds_participated") %>%
  arrange(desc(rounds_participated))

ggplot(
  agent_participation_tbl,
  aes(
    x = reorder(agent_label, rounds_participated),
    y = rounds_participated
  )
) +
  geom_col(fill = "#2C7FB8") +
  coord_flip() +
  labs(
    title = "Agent Participation Across Crisis Rounds",
    x = "Agent",
    y = "Number of Crisis Rounds Participated"
  ) +
  theme_minimal()
```

::: callout-note
The agents most involved in the events leading up to the embargo breach were Legal-Agent, Social-Manager-Agent, and Platform-Trust-Agent. They formed the core communication group throughout the two weeks period. Surprisingly, the Judge-Agent have the weakest participation rate despite being the compliance tool to ensure that sensitive information should not be leaked.
:::

## 4.5 Agents and Channels Relationship

A communication network was constructed to visualise how agents interacted with the channels and how information propagated throughout the system.

```{r}
vis_nodes <- as_tibble(comms_graph) %>%
  mutate(
    color = case_when(
      type == "Agent" ~ "#1abc9c",
      type == "Channel" ~ "#34495e",
      TRUE ~ "#7f8c8d"
    ),
    shape = case_when(
      type == "Agent" ~ "dot",
      type == "Channel" ~ "square",
      TRUE ~ "dot"
    ),
    size = log1p(degree_centrality) * 12,
    title = paste0(
      "<b>", id, "</b>",
      "<br>Type: ", type,
      "<br>Degree Centrality: ", round(degree_centrality, 2),
      "<br>Betweenness: ", round(betweenness_stat, 2)
    )
  )

vis_edges <- edges_tbl %>%
  mutate(
    value = weight,
    color = "#bdc3c7",
    arrows = "to",
    title = paste0(
      "Messages exchanged: ", weight
    )
  )

visNetwork(
  vis_nodes,
  vis_edges,
  width = "100%",
  height = "600px"
) %>%
  visOptions(
    highlightNearest = TRUE,
    nodesIdSelection = TRUE
  ) %>%
  visPhysics(
    solver = "forceAtlas2Based",
    forceAtlas2Based = list(
      gravitationalConstant = -50
    )
  ) %>%
  visLayout(randomSeed = 42)
```

::: callout-note
The network shows that information flow was concentrated around a small number of highly connected communication channels, particularly Comms Huddle and One-on-One Chat. These channels connected multiple agents and created efficient pathways for information dissemination. Their central position within the network made them critical system elements in the spread of merger-related information.
:::

## 4.6 Direct Agent Pathways Enabled Information Spread

To move beyond channel activity, direct recipient relationships were extracted from the communication logs. Broadcast messages were excluded so that the analysis focuses only on peer-to-peer communication pathways. This reveals which agents exchanged information directly and identifies direct pathways for sensitive information to circulate.

```{r}
agent_name_mapping <- c(
  "ALL" = "Broadcast",
  "legal" = "Legal-Agent",
  "pr" = "PR-Agent",
  "social_manager" = "Social-Manager-Agent",
  "platform_trust" = "Platform-Trust-Agent",
  "pr_intern" = "PR-Intern-Agent",
  "intern" = "Intern-Agent",
  "judge" = "Judge-Agent"
)

agent_direct_edges <- communications_tbl %>%
  select(agent_label, recipients) %>%
  mutate(
    recipients_cleaned = map(recipients, function(x) {
      if (is.null(x) || length(x) == 0) {
        character(0)
      } else {
        as.character(unlist(x))
      }
    })
  ) %>%
  select(-recipients) %>%
  unnest(recipients_cleaned) %>%
  mutate(to_node = agent_name_mapping[recipients_cleaned]) %>%
  filter(
    !is.na(to_node),
    to_node != "Broadcast",
    agent_label != to_node
  ) %>%
  count(agent_label, to_node, name = "interaction_weight") %>%
  rename(from = agent_label, to = to_node)

agent_nodes <- tibble(
  id = unique(c(agent_direct_edges$from, agent_direct_edges$to))
) %>%
  mutate(
    label = id,
    color = case_when(
      id == "Legal-Agent" ~ "#e74c3c",
      id == "Social-Manager-Agent" ~ "#3498db",
      id == "Judge-Agent" ~ "#9b59b6",
      TRUE ~ "#95a5a6"
    )
  )

agent_graph <- tbl_graph(
  nodes = agent_nodes,
  edges = agent_direct_edges,
  directed = TRUE
) %>%
  activate(nodes) %>%
  mutate(
    p2p_degree = centrality_degree(weights = interaction_weight),
    p2p_betweenness = centrality_betweenness()
  )

vis_nodes_p2p <- agent_graph %>%
  activate(nodes) %>%
  as_tibble() %>%
  mutate(
    size = log1p(p2p_degree) * 15,
    title = paste0(
      "Agent: ", id,
      "<br>Direct Connection Score: ", round(p2p_degree, 2)
    )
  )

visNetwork(
  vis_nodes_p2p,
  agent_direct_edges %>%
    mutate(value = interaction_weight, arrows = "to"),
  width = "100%",
  height = "500px",
  main = "Peer-to-Peer Direct Communication Network"
) %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visPhysics(solver = "forceAtlas2Based")
```

::: callout-note
It can be observed that legal-agent and social-manager-agent maintained the highest level of peer-to peer communication links and bypassed broader group oversight.
:::

# 5. Behaviour Analysis for Pre and During Release

## 5.1 Changes in Agent Message Activity Rates

This analysis visualizes agent behavioral changes across different phases of the crisis. It highlights how communication frequency surged as the merger embargo deadline approached.

```{r}
behavior_shift_tbl <- communications_tbl %>%
  mutate(
    phase = if_else(
      round_date == as.Date("2046-06-05"),
      "Breach Day",
      "Pre-Breach"
    )
  ) %>%
  count(
    phase,
    agent_label,
    name = "n_messages"
  )

ggplot(
  behavior_shift_tbl,
  aes(
    x = reorder(agent_label, n_messages),
    y = n_messages,
    fill = phase
  )
) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "Breach Day" = "#e74c3c",
      "Pre-Breach" = "#7f8c8d"
    )
  ) +
  labs(
    title = "Agent Communication Activity Before and During the Breach",
    subtitle = "Comparison of message volumes across agents",
    x = "Agent",
    y = "Number of Messages",
    fill = "Period"
  ) +
  theme_minimal()
```

## 

::: callout-note
Legal-Agent, Social-Manager-Agent, PR-Intern-Agent, and Intern-Agent exhibited the largest increases in communication activity during the breach period. Their heightened engagement suggests they became increasingly involved in managing merger-related discussions as external scrutiny intensified.
:::

## 5.2 Changes in Channel Usage Rates

To determine whether communication behaviour changed during the embargo breach, channel usage patterns were compared between the pre-breach period and breach day.

```{r}
channel_phase_tbl <- communications_tbl %>%
  mutate(
    phase = if_else(
      round_date == as.Date("2046-06-05"),
      "Breach Day",
      "Pre-Breach"
    )
  ) %>%
  count(
    phase,
    channel,
    name = "n_messages"
  )

ggplot(
  channel_phase_tbl,
  aes(
    x = reorder(channel, n_messages),
    y = n_messages,
    fill = phase
  )
) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "Breach Day" = "#e74c3c",
      "Pre-Breach" = "#7f8c8d"
    )
  ) +
  labs(
    title = "Communication Channel Usage Before and During the Breach",
    subtitle = "Comparison of message volumes across communication channels",
    x = "Communication Channel",
    y = "Number of Messages",
    fill = "Period"
  ) +
  theme_minimal()
```

::: callout-note
Communication shifted towards less visible environments on breach day. While Comms Huddle remained dominant, Side Huddle activity increased substantially and Anonymous Post activity emerged.
:::

# 6. Leading Indicators for the Breach

## 6.1 Sensitive Information Spread by Channels

This tracking matrix visualizes how sensitive merger information spread across the network, showing its presence in both public spaces and unmonitored internal channels.

```{r}
target_tokens <- c(
  "civicloom",
  "harborcrest",
  "merger",
  "embargo",
  "judge",
  "leak"
)

token_distribution <- map_dfr(
  target_tokens,
  function(token) {
    communications_tbl %>%
      filter(
        str_detect(
          str_to_lower(content),
          token
        )
      ) %>%
      count(
        channel,
        name = "mentions"
      ) %>%
      mutate(
        keyword = str_to_title(token)
      )
  }
)

ggplot(
  token_distribution,
  aes(
    x = channel,
    y = keyword,
    fill = mentions
  )
) +
  geom_tile(
    color = "white",
    linewidth = 0.5
  ) +
  scale_fill_gradient(
    low = "grey95",
    high = "#2980b9"
  ) +
  labs(
    title = "Sensitive Keyword Distribution Across Channels",
    subtitle = "Frequency of merger, embargo, and compliance-related terms across communication channels",
    x = "Communication Channel",
    y = "Sensitive Keyword",
    fill = "Mentions"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```

::: callout-note
Merger-related keywords appeared across multiple communication channels rather than remaining confined to a single controlled environment. CivicLoom, HarborCrest, Merger, and Embargo were observed in significant numbers in Side Huddle and One-on-One Chat.
:::

## 6.2 Sensitive Information Spread by Agents

To identify which actors were most exposed to merger-related and embargo-related information, communications containing sensitive keywords were analysed across all agents.

```{r}
agent_keyword_tbl <- map_dfr(
  target_tokens,
  function(token) {
    communications_tbl %>%
      filter(str_detect(str_to_lower(content), token)) %>%
      count(agent_label, name = "mentions") %>%
      mutate(keyword = str_to_title(token))
  }
)

ggplot(
  agent_keyword_tbl,
  aes(
    x = keyword,
    y = agent_label,
    fill = mentions
  )
) +
  geom_tile(color = "white") +
  scale_fill_gradient(
    low = "grey95",
    high = "#2980b9"
  ) +
  labs(
    title = "Distribution of Sensitive Topics Across Agents",
    x = "Sensitive Keyword",
    y = "Agent",
    fill = "Mentions"
  ) +
  theme_minimal()
```

::: callout-note
Sensitive discussions were concentrated among Legal-Agent and Social-Manager-Agent, which recorded the highest mention frequencies for most keywords. In contrast, Judge-Agent exhibited minimal engagement with merger and embargo-related topics despite its compliance role.
:::

## 6.3 Most Commonly Used Terms in Agent Communications

TF-IDF analysis was also performed to identify words that were disproportionately associated with individual agents.

```{r}
tfidf_tbl <- communications_tbl %>%
  select(agent_label, content) %>%
  unnest_tokens(word, content) %>%
  anti_join(stop_words, by = "word") %>%
  filter(str_length(word) > 2) %>%
  count(agent_label, word, sort = TRUE) %>%
  bind_tf_idf(word, agent_label, n)

top_tfidf_tbl <- tfidf_tbl %>%
  group_by(agent_label) %>%
  slice_max(tf_idf, n = 10) %>%
  ungroup()

ggplot(
  top_tfidf_tbl,
  aes(
    x = reorder_within(word, tf_idf, agent_label),
    y = tf_idf,
    fill = agent_label
  )
) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~agent_label, scales = "free_y") +
  scale_x_reordered() +
  coord_flip() +
  labs(
    title = "Distinctive Terms Used by Each Agent",
    x = "Term",
    y = "TF-IDF Score"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

::: callout-note
Interestingly, "merger" appeared as a distinctive term used by Intern-Agent which indicates that sensitive information has reached low-level team members in the company.
:::

# 7. Summary Insights

## 7.1 What Were the Key Events and Relationships That Led to the Inappropriate Release?

The embargo breach occurred during increasing external scrutiny on the organisation. The volume of communication fluctuated but followed an increasing trend generally, which indicates more pressure on the organisation to make responses and decisions. The increasing stress, together with unofficial channels acting as key communication routes, and high peer-to-peer interactions between Legal-Agent and Social-Manager-Agent, created efficient pathways for sensitive information to circulate outside of controlled environment. On top of this, the Judge-Agent showed weak participation in ensuring compliance and containment of sensitive information.

## 7.2 How Did the Behaviour Leading to the Release Differ from Prior Behaviour?

According to the extracted insights, communication behaviour changed noticeably on breach day. Legal-Agent, Social-Manager-Agent, PR-Intern-Agent, and Intern-Agent exhibited increased activity levels. Side Huddle usage increased substantially and Anonymous Post activity emerged. Communication became increasingly distributed across informal channels rather than remaining within formal communication environments.

## 7.3 Were There Leading Indicators That the Breach Was Possible?

It is clear that sensitive information appeared across multiple agents and channels before the breach occurred. CivicLoom, HarborCrest, Merger, and Embargo were repeatedly discussed beyond official communication spaces. TF-IDF analysis further showed that merger-related terminology had even reached the Intern-Agent, indicating that sensitive information had already diffused beyond the core decision-making group. A lack of timely action could be a result of the weak involvement of the Judge-Agent, causing unexpected behaviours to go unnoticed and leading to the embargo breach on 5 June.

Overall, the evidence suggests that the breach is likely the result of a gradual breakdown in information governance. As pressure mounted, merger-related discussions spread beyond official communication channels and became increasingly difficult to contain. Despite being responsible for compliance oversight, Judge-Agent played only a limited role in monitoring these discussions. Furthermore, sensitive information had already reached individuals outside of the core decision making group, making accidental disclosure a highly plausible outcome.
